### 2025-04-01 paper/paper.tex
draft paper

\documentclass[a4paper]{article}

% Packages
\usepackage{geometry}
\geometry{left=1.5cm, right=1.5cm, top=2.54cm, bottom=2.54cm}
\usepackage{graphicx, hyperref, setspace, amsmath, amssymb, titlesec, fancyhdr, multicol, parskip, indentfirst, etoolbox, caption, cite, hyperref, xcolor, svg}

% Title Formatting
\titleformat{\section}{\centering\large\scshape}{\thesection}{1em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{1em}{}
\setstretch{1.0} % Keep single spacing
\setlength{\parskip}{6pt} % Space between paragraphs
\titlespacing{\section}{0pt}{6pt}{6pt}
\titlespacing{\subsection}{0pt}{6pt}{6pt}
\titlespacing{\subsubsection}{0pt}{6pt}{6pt}
% Document Title
\title{
    \textbf{Draft v0 (I need to migrate my citation data to latex which is still a mess right now)} 
}

\date{} % No date
\captionsetup{labelfont={small,sc}, textfont={small,sc}}
% Section Numbering
% Define numbering format
\renewcommand{\thesection}{\Roman{section}.}
\renewcommand{\thesubsection}{\textit{\Alph{subsection}.}}
\renewcommand{\thesubsubsection}{\textit{\arabic{subsubsection}.}}
\renewcommand{\thetable}{\Roman{table}} % Set table numbering to Roman
\renewcommand{\thefigure}{\Roman{figure}} % Number figures in Roman numerals

% Make titles italic as well

\titleformat{\subsection}{\normalfont\large\itshape}{\thesubsection}{1em}{}
\titleformat{\subsubsection}{\normalfont\itshape}{\thesubsubsection}{1em}{}

\fancyfoot[L]{\includegraphics[width=1.5cm]{cc-by.png}}  % Adjust path & size
\fancyfoot[R]{\includegraphics[width=1.5cm]{cc-by.png}}  % Adjust path & size
\fancyfoot[C]{This work is licensed under a Creative Commons Attribution 4.0 International License.\\ \thepage}

\begin{document}

\maketitle
\vspace{-1.5cm}

% Authors Block
\begin{multicols}{3}
    \centering
    \textbf{Yang Li}\\
    \textit{University of the West of England}\\
    \textit{yang4.li@uwe.ac.uk}\\
    \vfill

    \textbf{Marco Perez Hernandez}\\
    \textit{University of the West of England}\\
    \textit{marco.perezhernandez@uwe.ac.uk}\\
    \vfill

    \textbf{Mehmet Aydin}\\
    \textit{University of the West of England}\\
    \textit{mehmet.aydin@uwe.ac.uk}\\
    \vfill
\end{multicols}

\singlespacing
\setlength{\parskip}{6pt}
\setlength{\parindent}{0.5cm}

\begin{abstract}
The abstract should be written after the rest of the paper is completed.
\end{abstract}

\begin{multicols}{2}
\setlength{\columnsep}{0.5cm}

\section{Introduction}
Plenty of research has been done on multi-agent reinforcement learning (MARL) in fully cooperative games. There is currently a gap in the non-cooperative setting. In fully cooperative games, there is no conflict between the agents. If agents can communicate, they are motivated to form a firm alliance that essentially works a collective mind which is the one and only conscious being. Thus, in a game theory sense, a fully cooperative game where agents share information is equivalent to a single agent game where the collective agent tries to find the optimal joint action, given the joint observation. To do that in a traditional single agent setting where the environment is stationary and non-adversarial, one only needs to learn the environment mechanics (the state transition function) and the reward function so that the Q-values of an action in a given state can be computed or predicted by a function approximator. A Pareto optimal pure (deterministic) strategy is guaranteed to exist in fully cooperative finite games, meaning that the agents can always deterministically choose the optimal action that has the highest payoff. The only thing that keeps the agents from forming a cluster in practice is the hardware incapability (computational power, real-world constraints, etc.).

Non-cooperative (at least partially competitive) environments, however, involves conflicts that make learning more complicated. Agents might belong to different parties that would like to have private control to protect their own interests. For example, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss, if we omit the long term economic growth.

In fully cooperative games, we can always say that 5 rewards is better than 2 rewards, but what about the rewards $\langle 1,5 \rangle$ and  $\langle 10,1 \rangle$ where each agent gets one value in the vector? Rational self-interested agents will have to fight for what is best for them. Many solutions in game theory are possible. One of the commonly accepted solutions for is the Nash equilibrium (NE). NEs are normally what we want the agents to converge to because convergence means stability and strict NEs are the only stable points in the game. A finite perfect-information game always has a pure NE if the game is asynchronous (players take turns), such as Chess, checkers and Go. Algorithms like minimax tree search can be applied to those games to find NEs. In some cases, there may only be a mixed NE where players adopt randomized strategies consisting of probability distributions over the action space.

Currently, most SOTA MARL algorithms cannot learn a mixed policy. Largely due to the studied games are mostly fully cooperative or asynchronous. As we will show in the experiments, those algorithms will fail to find the optimal action in games specifically designed to challenge the agents' competitive thinking. We need to reconsider our approach if we want to build agents that are capable of dealing with non-fully-cooperative environments.

\subsection{Aim}
The aim of this study is to point out that current SOTA algorithms can be misleadingly performing well in non-fully-cooperative environments by both formal analysis and showing specific game scenarios that are designed to challenge the agents from a game theory perspective.

\subsection{Background}

A stochastic game is formally defined by a tuple:

$G = \langle I,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
\begin{itemize}
    \item $I$: the set of agents, from 1 to \(n\).
    \item $S$: the game’s state space.
    \item $S_0$: the set of possible initial states.
    \item $S_t$: the set of terminal states.
    \item $\{A_1,\ldots,A_n\}$: one action space for each agent.

    Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent.

    $R_i:S \times A^n \times S \rightarrow \mathbb{R}$
    \item $T$: the state transition function.

    $T:S \times A^n \times S \rightarrow [0,1]$

    It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
    \begin{equation}
    \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
    \label{eq:statetransition}
    \end{equation}
    It is convenient to define $T^*:S \times A^n \rightarrow \Pi_S$, where $\Pi_S$ denotes the set of all probability distribution over the set S.

    Some games incorporate partial observability and define an observations space, which is irrelevant to the purpose of this study.
\end{itemize}

 Starting from a randomly chosen $s_0 \in S_0$, at each timestep $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$. The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent. If $s_t \in S_t$, the game ends. Otherwise, the game continues in the next timestep.

The goal is to find the optimal strategy (policy) for an agent to pick actions given the current state information they receive from the environment, such that their expected reward is maximized.

To demonstrate the problem with competitive games, let's analyze some simple normal form games. A normal form game has only two states, the starting state and the end state. The game ends immediately after all agents pick an action. Thus, they can be represented using the reward matrix. Consider rock-paper-scissors (simple one-shot matrix game) in Table~\ref{table:rps} for example. Each entry represents the reward for the row player and the column player, respectively.

\noindent
\begin{minipage}{\columnwidth}
\captionof{table}{Rock Paper Scissors} 
\label{table:rps}
\centering
\begin{tabular}{c|ccc}
    & r & p & s\\
    \hline
    r & 0,0 & -1,1 & 1,-1 \\
    p & 1,-1 & 0,0 & -1,1 \\
    s & -1,1 & 1,-1 & 0,0 \\
\end{tabular}
\end{minipage}

The only NE in the game is that two players play each of the three actions with 1/3 chance. If any player deviates from the NE, their expected reward does not increase, and the other player can exploit them by playing the action that counters their most frequent action. Here is a more complicated game called biased rock paper scissors.

\noindent
\begin{minipage}{\columnwidth}
\captionof{table}{Biased Rock Paper Scissors} 
\label{table:biasedrps}
\centering
\begin{tabular}{c|ccc}
    & r & p & s\\
    \hline
    r & 50,50 & 25,75 & 100,0 \\
    p & 75,25 & 50,50 & 45,55 \\
    s & 0,100 & 55,45 & 50,50 \\
\end{tabular}
\end{minipage}

The only NE in the game is that both players adopt the following policy.
\begin{equation}
\pi=(P(r)=0.0625,P(p)=0.6250,P(s)=0.3125)
\label{eq:biasedrpspolicy}
\end{equation}

The RL algorithm needs to learn $\pi$, not the optimal action because there is no optimal action. For competitive play after the learning process, the agent should play randomly according to the learned distribution.

Algorithms that are carried over from the traditional single agent RL domain, such as the Q learning algorithm family, do not incorporate randomized policy. Thus, they cannot learn $\pi$, even though they may behave randomly during the learning process (using an $\epsilon$-greedy exploration vs. exploitation strategy, for example). They learn the Q-value of each action under each state, which is supposed to represent the payoff of that action. They choose the action with the best payoff when they are not exploring. There are research addressing this problem using a manually calculated distribution over the actions based on the payoff, such that the higher the payoff of an action, the more likely the action will be chosen. This approach intuitively makes sense. The problem is that in any NE, for any action that are chosen with a positive probability, the payoff of the actions are the same. If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions have the same payoff.

Suppose the algorithm can converge to a NE, then all Q-values should be the same (ignoring worthless actions). Such an algorithm will output a uniform distribution. The algorithm is essentially learning a pure random policy. This contradicts with the assumption that they converge to a NE. We have shown that approach cannot work through a proof by contradiction.

Most MARL problems studied are much more than single state matrix games and have astronomically large state/action spaces. It is impossible for the agents to converge to a global NE in the first place because of the limited computational power. Because of the complexity of the environment that many studies carry out in, picking the action with the best Q-value still gives reasonably good results since at least the agents learned to interact with the environment. The competitiveness of the game is not going to play a huge role before the agents can really predict the state transitions.

\section{Methods}
Here I will define the mechanics of the game Food Chain (this is now the official name). I also will have to migrate the tables and charts from my previous Word documents to explain the unique characteristics that make it challenging which are not found in other games (especially compare to the hide and seek game).

The game is complex enough to eliminate any tabular approaches. The state space is roughly 1060 for just a 9x9 board and can grow exponentially. Function approximation methods, such as deep neural networks or Monte Carlo methos must be used to solve the whole game. The spacial information on the game board can also be a good challenge to algorithms using CNNs. The game is still simple enough to be human playable. The mechanics of the game (the state transition function) is very straightforward and easy for the agents to learn, so that they can focus on agent interactions. Because of the large state space, it is impossible for us to check whether the agents converged to a global NE since it is NP-Hard. Yet it has carefully designed mechanics (see my Casual Monthly Progress Report for Nov 2024) which very often force the agents to adopt random policies (and addressing the coordination problem, too). These are the leaf nodes of the game tree, which is equivalent to a simple matrix game. We can actually check whether the agents converge to NEs in those sub-cases easily. So, in summary.
\begin{itemize}
    \item It is simple enough to let the agents focus on interactions instead of the environment.
    \item It is complex enough to be interesting.
    \item It forces the agents to constantly address the mixed strategy problem.
    \item We can analyze the performance of the agents with game theory.
\end{itemize}

\section{Results}
Figure \ref{fig:hopperreward}, \ref{fig:frogreward}, \ref{fig:snakereward} shows the average test return for each type of agent after training in a randomly generated game environment. Figure \ref{fig:scenhopperreward}, \ref{fig:scenfrogreward} demonstrates that in the carefully set up matching pennies scenario. The results show that the algorithms could not reach the optimal reward which can be computed using linear algebra (which is also another strong aspect of the game. we can analyze the game formally and see whether the agents reached optimality). More formal analysis coming...

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_global_hopper.svg}
\captionof{figure}{hopper rewards}
\label{fig:hopperreward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_global_frog.svg}
\captionof{figure}{frog rewards}
\label{fig:frogreward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_global_snake.svg}
\captionof{figure}{snake rewards}
\label{fig:snakereward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_scen_hopper.svg}
\captionof{figure}{snake rewards}
\label{fig:scenhopperreward}
\end{minipage}

\noindent
\begin{minipage}{\columnwidth}
\centering
\includesvg[width=1\textwidth]{plots/rewards_scen_frog.svg}
\captionof{figure}{snake rewards}
\label{fig:scenfrogreward}
\end{minipage}

\section{Discussion}

\section{Conclusion}

\section*{Funding}
This research was funded by the University of the West of England.

\begin{thebibliography}

\end{thebibliography}

\end{multicols}
\end{document}

### 2025-04-15 paper/paper.tex
v1

@@
 \geometry{left=1.5cm, right=1.5cm, top=2.54cm, bottom=2.54cm}
-\usepackage{graphicx, hyperref, setspace, amsmath, amssymb, titlesec, fancyhdr, multicol, parskip, indentfirst, etoolbox, caption, cite, hyperref, xcolor, svg}
+\usepackage{graphicx, hyperref, setspace, amsmath, amssymb, titlesec, multicol, parskip, indentfirst, etoolbox, caption, xcolor, svg}
+\usepackage[style=ieee]{biblatex}
+% Import the bibliography file
+\addbibresource{ref.bib}
 
@@
 \title{
-    \textbf{Draft v0 (I need to migrate my citation data to latex which is still a mess right now)} 
+    \textbf{Draft v1}
 }
 
-\date{} % No date
+\date{}
 \captionsetup{labelfont={small,sc}, textfont={small,sc}}
@@
 
-% Make titles italic as well
-
+% Make subsection titles italic as well
 \titleformat{\subsection}{\normalfont\large\itshape}{\thesubsection}{1em}{}
 \titleformat{\subsubsection}{\normalfont\itshape}{\thesubsubsection}{1em}{}
-
-\fancyfoot[L]{\includegraphics[width=1.5cm]{cc-by.png}}  % Adjust path & size
-\fancyfoot[R]{\includegraphics[width=1.5cm]{cc-by.png}}  % Adjust path & size
-\fancyfoot[C]{This work is licensed under a Creative Commons Attribution 4.0 International License.\\ \thepage}
 
@@
 \section{Introduction}
+Many real-world problems can be modeled as non-cooperative games, such as business competition, political campaigns, stock trading, bidding in auctions, cyber security, and traffic routing.
+These problems feature the competitive nature of games because they share one thing in common, parties with conflicting interests.
 Plenty of research has been done on multi-agent reinforcement learning (MARL) in fully cooperative games.
-There is currently a gap in the non-cooperative setting.
-In fully cooperative games, there is no conflict between the agents.
-If agents can communicate, they are motivated to form a firm alliance that essentially works a collective mind which is the one and only conscious being.
+There is currently a gap in the non-cooperative setting \cite{zhu2024survey}.
+Most current SOTA MARL algorithms are designed for cooperative games.
+
+In fully cooperative games, there is no conflict between agents.
+If agents can communicate, they are motivated to form a firm alliance that essentially works as a collective mind that is the one and only conscious being.
 Thus, in a game theory sense, a fully cooperative game where agents share information is equivalent to a single agent game where the collective agent tries to find the optimal joint action, given the joint observation.
@@
 
-Non-cooperative (at least partially competitive) environments, however, involves conflicts that make learning more complicated.
-Agents might belong to different parties that would like to have private control to protect their own interests.
-For example, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss, if we omit the long term economic growth.
+Non-cooperative environments, however, involve conflicts that make learning more complicated.
+Agents may belong to different parties that would like to have private control to protect their own interests.
+For example, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss, if we omit the long-term economic growth.
 
-In fully cooperative games, we can always say that 5 rewards is better than 2 rewards, but what about the rewards $\langle 1,5 \rangle$ and  $\langle 10,1 \rangle$ where each agent gets one value in the vector?
+For cooperative games, we can always say that 5 rewards is better than 2 rewards, but what about the rewards $\langle 1,5 \rangle$ and  $\langle 10,1 \rangle$ where each agent gets one value in the vector?
 Rational self-interested agents will have to fight for what is best for them.
@@
 NEs are normally what we want the agents to converge to because convergence means stability and strict NEs are the only stable points in the game.
-A finite perfect-information game always has a pure NE if the game is asynchronous (players take turns), such as Chess, checkers and Go.
-Algorithms like minimax tree search can be applied to those games to find NEs.
-In some cases, there may only be a mixed NE where players adopt randomized strategies consisting of probability distributions over the action space.
+Generally, there may only be a mixed NE, meaning that players adopt randomized strategies consisting of probability distributions over the action space.
 
@@
 
-\subsection{Background}
+\subsection{Stochastic Game}
 
@@
 
-The goal is to find the optimal strategy (policy) for an agent to pick actions given the current state information they receive from the environment, such that their expected reward is maximized.
+The goal is to find the optimal strategy (policy) for an agent to pick actions given the current state information they receive from the environment, such that their expected cumulative reward is maximized.
 
@@
 
-Algorithms that are carried over from the traditional single agent RL domain, such as the Q learning algorithm family, do not incorporate randomized policy.
-Thus, they cannot learn $\pi$, even though they may behave randomly during the learning process (using an $\epsilon$-greedy exploration vs.
+\section{MARL in Competitive Games}
+Algorithms that are carried over from the traditional single agent RL domain, such as Q-learning, do not incorporate randomized policy, even though they may behave randomly during the learning process (using an $\epsilon$-greedy exploration vs.
 exploitation strategy, for example).
-They learn the Q-value of each action under each state, which is supposed to represent the payoff of that action.
-They choose the action with the best payoff when they are not exploring.
-There are research addressing this problem using a manually calculated distribution over the actions based on the payoff, such that the higher the payoff of an action, the more likely the action will be chosen.
+In the single agent setting, the agent only needs to learn the expected long-term return (the Q-value) of each action and deterministically pick the highest.
+Thus, they cannot learn $\pi$.
+
+There is a family of algorithms called Joint Action Learning (JAL) that addresses the problem by incorporating the Q-values for the joint action.
+Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games.
+It computes local minimax solutions for each state based on the learned joint Q-values.
+Nash Q-learning \cite{hu2003nash} is a similar algorithm that computes the local NE instead.
+It converges to a NE under highly strict assumptions that is found in almost no games.
+Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees.
+\cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games.
+
+Some other research \cite{qu2020distributed} tried to address this problem using a manually calculated distribution over the actions based on the payoff, such that the higher the payoff of an action, the more likely the action will be chosen.
 This approach intuitively makes sense.
@@
 The competitiveness of the game is not going to play a huge role before the agents can really predict the state transitions.
+Currently, many fully competitive games are studied in the MARL community such as Pommerman \cite{resnick2018pommerman}, Go \cite{silver2017mastering}, StarCraft II \cite{vinyals2019grandmaster}, and Dota 2 \cite{berner2019dota}.
+There are also some general environments that are non-cooperative such as hide-and-seek \cite{baker2019emergent}, Neural MMO \cite{suarez2021neural}, PettingZoo \cite{terry2021pettingzoo}, and Melting Pot \cite{leibo2021scalable}.
+However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games.
+FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.
 
 \section{Methods}
-Here I will define the mechanics of the game Food Chain (this is now the official name).
+We define the mechanics of the game Food Chain.
 I also will have to migrate the tables and charts from my previous Word documents to explain the unique characteristics that make it challenging which are not found in other games (especially compare to the hide and seek game).
@@
 
-\begin{thebibliography}
-
-\end{thebibliography}
+% Prints bibliography
+\printbibliography
 

### 2025-05-13 paper/paper.tex
paper v2

@@
 \title{
-    \textbf{Draft v1}
+    \textbf{Draft v2}
 }
@@
 \date{}
-\captionsetup{labelfont={small,sc}, textfont={small,sc}}
+\captionsetup{labelfont={small,sc}, textfont={small,sc}, hypcap=false}
+%\captionsetup{labelfont={small,sc}, textfont={small,sc}}
 % Section Numbering
@@
 \section{Introduction}
-Many real-world problems can be modeled as non-cooperative games, such as business competition, political campaigns, stock trading, bidding in auctions, cyber security, and traffic routing.
+Many real-world problems can be modelled as non-cooperative games, such as business competition, political campaigns, stock trading, bidding in auctions, cyber security, and traffic routing.
 These problems feature the competitive nature of games because they share one thing in common, parties with conflicting interests.
@@
 Agents may belong to different parties that would like to have private control to protect their own interests.
-For example, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss, if we omit the long-term economic growth.
+For example, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss, if we omit the long-term economic growth, especially when we consider options, futures, etc.
+Another example is cyber security, which must study the conflicts between attackers and defenders.
 
@@
 Many solutions in game theory are possible.
-One of the commonly accepted solutions for is the Nash equilibrium (NE).
+One of the commonly accepted solutions is the Nash equilibrium (NE).
+NE describes a situation where no player can benefit by deviating from their current strategy independently.
 NEs are normally what we want the agents to converge to because convergence means stability and strict NEs are the only stable points in the game.
@@
 \subsection{Aim}
-The aim of this study is to point out that current SOTA algorithms can be misleadingly performing well in non-fully-cooperative environments by both formal analysis and showing specific game scenarios that are designed to challenge the agents from a game theory perspective.
+The aim of this study is to point out that current SOTA algorithms can be misleadingly performing well in non-fully-cooperative environments.
+Furthermore, we will find out the conditions a game must satisfy for the algorithms to work by both formal analysis and specific game scenarios that are designed to challenge the agents from a game theory perspective.
 
@@
 
-A stochastic game is formally defined by a tuple:
+A (finite) stochastic game is formally defined by a tuple:
 
@@
 \begin{itemize}
-    \item $I$: the set of agents, from 1 to \(n\).
+    \item $I$: the set of agents, numbered from 1 to $n$.
     \item $S$: the game’s state space.
@@
     \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
-    \label{eq:statetransition}
+    \label{eq:state_transition}
     \end{equation}
-    It is convenient to define $T^*:S \times A^n \rightarrow \Pi_S$, where $\Pi_S$ denotes the set of all probability distribution over the set S.
 
@@
 
-The goal is to find the optimal strategy (policy) for an agent to pick actions given the current state information they receive from the environment, such that their expected cumulative reward is maximized.
+The way a player plays the game defines a policy (strategy) as a function that computes the probability of selecting each action given the current state, $\pi:S \times A^n \rightarrow [0,1]$.
+A strategy is said to be pure if it is deterministic (puts all the probability on a single action).
+Otherwise, we call it a mixed strategy.
+A set of strategies, one for each player, is called a profile $\pi^n$.
+Given the profile, we can compute the expected reward for each player in the game.
+The goal of MARL is to find the optimal strategy for an agent, or even the optimal profile, such that their expected cumulative reward is maximized.
 
-To demonstrate the problem with competitive games, let's analyze some simple normal form games.
-A normal form game has only two states, the starting state and the end state.
+\section{MARL in Non-cooperative Games}
+John Nash is famous for his study on non-cooperative Games and his concept of Nash Equilibrium (NE).
+A NE is a profile where none shall increase their expected reward by shifting to another policy alone.
+Formally, let $R_i^*(\pi^n)$ be the function to compute the expected reward of the i-th player given the profile.
+Let $\pi_i$ be a policy of the i-th player and $\Pi_i$ be the policy space of the i-th player.
+Let $\pi^{-i}$ denote $\pi^n - \{\pi_i\}$, the set of strategies of all other players.
+A profile $\pi^n$ is a NE if
+\begin{equation}
+\forall i \in \{1, ..., n\}, \forall \pi_i \in \Pi_i: R_i^*(\{\pi_i\} \cup \pi^{-i}) \leqslant R_i^*(\pi^n)
+\label{eq:nash}
+\end{equation}
+A NE is called strict if the $\leqslant$ is replaced by $<$.
+
+A game is finite if it involves a finite number of players where
+the action spaces and the state space are finite, and the game always ends in a finite number of turns.
+John Nash proved that every finite game has at least one NE.
+To solve a game in the context of this paper means to find at least one NE.
+Games that are not finite can be ill-conditioned and have no NE.
+For example, a game where the person who names the greatest number wins has no best strategy.
+For the rest of the paper, games refer to finite games.
+
+To demonstrate the problem with competitive games, let's analyse some simple normal form games.
+A normal form game has only two states, the starting state $S_0$ and the end state $S_1$.
 The game ends immediately after all agents pick an action.
 Thus, they can be represented using the reward matrix.
-Consider rock-paper-scissors (simple one-shot matrix game) in Table~\ref{table:rps} for example.
+Consider the simple one-shot matrix game rock-paper-scissors in Table \ref{table:rps} for example.
 Each entry represents the reward for the row player and the column player, respectively.
@@
 \captionof{table}{Biased Rock Paper Scissors} 
-\label{table:biasedrps}
+\label{table:biased_rps}
 \centering
@@
 \pi=(P(r)=0.0625,P(p)=0.6250,P(s)=0.3125)
-\label{eq:biasedrpspolicy}
+\label{eq:biased_rps_policy}
 \end{equation}
@@
 
-\section{MARL in Competitive Games}
-Algorithms that are carried over from the traditional single agent RL domain, such as Q-learning, do not incorporate randomized policy, even though they may behave randomly during the learning process (using an $\epsilon$-greedy exploration vs.
+Independent Action Learning (IAL) algorithms that are carried over from the traditional single agent RL domain, such as Q-learning, learns the expected long-term reward for each action in each state.
+In the single agent setting, the agent only needs to learn the expected long-term return (the Q-value) of each action and deterministically pick the highest.
+They do not incorporate randomized policy, even though they may behave randomly during the learning process (using an $\epsilon$-greedy exploration vs.
 exploitation strategy, for example).
-In the single agent setting, the agent only needs to learn the expected long-term return (the Q-value) of each action and deterministically pick the highest.
 Thus, they cannot learn $\pi$.
+This is the mixed strategy problem.
 
-There is a family of algorithms called Joint Action Learning (JAL) that addresses the problem by incorporating the Q-values for the joint action.
+The gradient methods approach works around this issue by learning the policies directly, without putting an estimation on the individual action reward.
+There is also a family of algorithms called Joint Action Learning (JAL) which addresses the problem by incorporating the values for the joint action.
 Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games.
@@
 \cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games.
+Both gradient methods and JAL are centralized learning which needs to know the action of other agents.
+In addition, gradient methods require full knowledge of the reward function.
 
+Some study addresses this issue by picking the action with the highest empirical return \cite{bjornsson2009cadiaplayer}, and received reasonably good outcome.
 Some other research \cite{qu2020distributed} tried to address this problem using a manually calculated distribution over the actions based on the payoff, such that the higher the payoff of an action, the more likely the action will be chosen.
-This approach intuitively makes sense.
+The latter approach is more sophisticated and intuitively makes sense.
 The problem is that in any NE, for any action that are chosen with a positive probability, the payoff of the actions are the same.
-If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions have the same payoff.
-
-Suppose the algorithm can converge to a NE, then all Q-values should be the same (ignoring worthless actions).
-Such an algorithm will output a uniform distribution.
-The algorithm is essentially learning a pure random policy.
-This contradicts with the assumption that they converge to a NE.
-We have shown that approach cannot work through a proof by contradiction.
+We will take a in-depth look in the formal analysis section.
 
@@
 
+\section{Formal Analysis}
+In this part of the paper, we show the feasibility of single-action learning (SAL) algorithms in non-cooperative games with formal analysis.
+We demonstrate the necessary conditions the game must satisfy in order to apply the IAL algorithms.
+
+In fact, finite games can be translated into normal form games.
+If we encapsulate the following states as a subgame, we end up with a reward matrix for the initial state with the rewards being the expected rewards of the subgames.
+It will be simpler if we focus on normal form games for the formal part.
+Since they only have one state $S_0$ for agents to take actions, we simply write $\pi(a)$ instead of $\pi(S_0, a)$ to denote the probability of taking action $a$.
+
+A direct consequence of (\ref{eq:nash}) is that at any NE $\pi^n$, given the policy set of the other players $\pi^{i-1}$, player $i$ is indifferent from all actions that are chosen with a positive probability.
+This means that the actions are split into two sets, the worthy actions that are chosen with a positive probability and the worthless ones that will never be piked.
+For example, we can add a suicidal action in the rock-paper-scissors game where the player loses once executing it.
+Such an action should never be selected by the players.
+Additionally, the expected reward will be the same for every worthy action.
+In other words, the player can choose the worthy actions arbitrarily, given that other players commit to the NE.
+Many find this fact about NE counter-intuitive.
+If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions indeed have the same payoff.
+
+The above fact can be formally stated as the following.
+Assume $\pi^n$ is a NE.
+Let $Q_i(a_j, \pi^{-i})$ be the expected reward of action $a_j \in A_i$ for agent $i$, given the rest of the profile $\pi^{-i}$.
+Let $A^+_i$ denote the set of worthy actions of agent $i$, namely $A^+_i = \{a | a \in A_i \wedge \pi_i(a) > 0 \}$.
+We have:
+\begin{equation}
+\begin{aligned}
+\forall a_j, a_k \in A^+_i, Q_i(a_j, \pi^{-i}) = Q_i(a_k, \pi^{-i})
+\label{eq:nash_indifferent}
+\end{aligned}
+\end{equation}
+
+Suppose a IAL algorithm can converge to a NE, then all worthy action's Q-values must converge to the same for an agent.
+It does not matter how the algorithm computes a probability distribution over the actions at the end, it cannot make any discriminations among those worthy actions based on equal statistics.
+Such an algorithm will output at best a uniform distribution over the worthy actions at a NE.
+This contradicts with the assumption that they converge to a NE unless the NE is full of uniformly random policies which is not the case in most games.
+We have shown that pure IAL approaches cannot work through proof by contradiction.
+
+This gives us a necessary condition for IAL algorithms to work.
+IAL algorithms can converge to a NE only if the NE consists of uniformly random policies over the set of all worthy actions.
+Note that this also includes any pure policies (agents deterministically select one action).
+It can be formally stated as follows.
+\begin{equation}
+\exists \pi^n, \forall \pi_i \in \pi^n, \forall a_j \in A_i, \pi_i(a_j, \pi^{-i}) = Q_i(a_k, \pi^{-i})
+\label{eq:necessary_condition}
+\end{equation}
+Because in cooperative games, there is no need to use mixed strategies, the NEs are essentially pure.
+This means all cooperative games satisfy (\ref{eq:necessary_condition}), which explains why IAL works in cooperative games.
+
 \section{Methods}
+\(------Under-Construction-Line------\)
+
+To verify our formal analysis, we test the algorithms in a game that can turn on and off the necessary conditions described in the previous section.
 We define the mechanics of the game Food Chain.
@@
 The state space is roughly 1060 for just a 9x9 board and can grow exponentially.
-Function approximation methods, such as deep neural networks or Monte Carlo methos must be used to solve the whole game.
+Function approximation methods, such as deep neural networks or Monte Carlo methods must be used to solve the whole game.
 The spacial information on the game board can also be a good challenge to algorithms using CNNs.
@@
 So, in summary.
+
 \begin{itemize}
@@
     \item It forces the agents to constantly address the mixed strategy problem.
-    \item We can analyze the performance of the agents with game theory.
+    \item We can analyse the performance of the agents with game theory.
 \end{itemize}
 
+\subsection{Biased Matching Pennies}
+This scenario forces the agents to address the mixed strategy problem. We set up the environment as follows.
+
+\noindent
+\begin{minipage}{\columnwidth}
+\centering
+\includegraphics[width=0.4\textwidth]{images/biased_matching_pennies.png}
+\captionof{figure}{Biased Matching Pennies}
+\label{fig:biased_matching_pennies}
+\end{minipage}
+
+\subsection{Tree Maze Chase}
+This scenario satisfies (\ref{eq:necessary_condition}). It essentially turns off the mixed strategy problem. We set up the game environment as follows.
+
+\noindent
+\begin{minipage}{\columnwidth}
+\centering
+\includegraphics[width=0.8\textwidth]{images/tree_maze_chase.png}
+\captionof{figure}{Tree Maze Chase}
+\label{fig:tree_maze_chase}
+\end{minipage}
+
 \section{Results}
-Figure \ref{fig:hopperreward}, \ref{fig:frogreward}, \ref{fig:snakereward} shows the average test return for each type of agent after training in a randomly generated game environment.
-Figure \ref{fig:scenhopperreward}, \ref{fig:scenfrogreward} demonstrates that in the carefully set up matching pennies scenario.
+Figure \ref{fig:hopper_reward}, \ref{fig:frog_reward}, \ref{fig:snake_reward} shows the average test return for each type of agent after training in a randomly generated game environment.
+Figure \ref{fig:biased_matching_pennies_hopper_reward}, \ref{fig:biased_matching_pennies_frog_reward} demonstrates that in the carefully set up matching pennies scenario.
 The results show that the algorithms could not reach the optimal reward which can be computed using linear algebra (which is also another strong aspect of the game.
-we can analyze the game formally and see whether the agents reached optimality).
-More formal analysis coming...
+we can analyse the game formally and see whether the agents reached optimality).
 
@@
 \captionof{figure}{hopper rewards}
-\label{fig:hopperreward}
+\label{fig:hopper_reward}
 \end{minipage}
@@
 \captionof{figure}{frog rewards}
-\label{fig:frogreward}
+\label{fig:frog_reward}
 \end{minipage}
@@
 \captionof{figure}{snake rewards}
-\label{fig:snakereward}
+\label{fig:snake_reward}
 \end{minipage}
@@
 \includesvg[width=1\textwidth]{plots/rewards_scen_hopper.svg}
-\captionof{figure}{snake rewards}
-\label{fig:scenhopperreward}
+\captionof{figure}{biased matching pennies hopper rewards}
+\label{fig:biased_matching_pennies_hopper_reward}
 \end{minipage}
@@
 \includesvg[width=1\textwidth]{plots/rewards_scen_frog.svg}
-\captionof{figure}{snake rewards}
-\label{fig:scenfrogreward}
+\captionof{figure}{biased matching pennies frog rewards}
+\label{fig:biased_matching_pennies_frog_reward}
 \end{minipage}

### 2025-06-04 paper/paper.tex
paper v6

--- 2025-04-15 paper/paper.tex
@@
-\documentclass[a4paper]{article}
+\documentclass[]{interact}
 
-% Packages
-\usepackage{geometry}
-\geometry{left=1.5cm, right=1.5cm, top=2.54cm, bottom=2.54cm}
-\usepackage{graphicx, hyperref, setspace, amsmath, amssymb, titlesec, multicol, parskip, indentfirst, etoolbox, caption, xcolor, svg}
-\usepackage[style=ieee]{biblatex}
-% Import the bibliography file
-\addbibresource{ref.bib}
+\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
+\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
+%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
+%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
+%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled
+\usepackage{svg}
 
-% Title Formatting
-\titleformat{\section}{\centering\large\scshape}{\thesection}{1em}{}
-\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{1em}{}
-\setstretch{1.0} % Keep single spacing
-\setlength{\parskip}{6pt} % Space between paragraphs
-\titlespacing{\section}{0pt}{6pt}{6pt}
-\titlespacing{\subsection}{0pt}{6pt}{6pt}
-\titlespacing{\subsubsection}{0pt}{6pt}{6pt}
-% Document Title
-\title{
-    \textbf{Draft v1}
-}
+\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
+\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
+\makeatletter% @ becomes a letter
+\def\NAT@def@citea{\def\@citea{\NAT@separator}}% Suppress spaces between citations using natbib.sty
+\makeatother% @ becomes a symbol again
 
-\date{}
-\captionsetup{labelfont={small,sc}, textfont={small,sc}}
-% Section Numbering
-% Define numbering format
-\renewcommand{\thesection}{\Roman{section}.}
-\renewcommand{\thesubsection}{\textit{\Alph{subsection}.}}
-\renewcommand{\thesubsubsection}{\textit{\arabic{subsubsection}.}}
-\renewcommand{\thetable}{\Roman{table}} % Set table numbering to Roman
-\renewcommand{\thefigure}{\Roman{figure}} % Number figures in Roman numerals
+\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
 
-% Make subsection titles italic as well
-\titleformat{\subsection}{\normalfont\large\itshape}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}{\normalfont\itshape}{\thesubsubsection}{1em}{}
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
 
@@
 
+%\articletype{ARTICLE TEMPLATE} Specify the article type or omit as appropriate
+
+\title{Behavioural Analysis of Independent Value-Based Learning in Non-cooperative Games}
+
+\author{
+\name{Yang Li\textsuperscript{a}\thanks{CONTACT Yang Li. Email: yang4.li@uwe.ac.uk}, Marco Perez Hernandez\textsuperscript{a} and Mehmet Emin Aydin\textsuperscript{a}}
+\affil{\textsuperscript{a} University of the West of England, Coldharbour Lane, Frenchay Campus, Bristol, UK}
+}
+
 \maketitle
-\vspace{-1.5cm}
-
-% Authors Block
-\begin{multicols}{3}
-    \centering
-    \textbf{Yang Li}\\
-    \textit{University of the West of England}\\
-    \textit{yang4.li@uwe.ac.uk}\\
-    \vfill
-
-    \textbf{Marco Perez Hernandez}\\
-    \textit{University of the West of England}\\
-    \textit{marco.perezhernandez@uwe.ac.uk}\\
-    \vfill
-
-    \textbf{Mehmet Aydin}\\
-    \textit{University of the West of England}\\
-    \textit{mehmet.aydin@uwe.ac.uk}\\
-    \vfill
-\end{multicols}
-
-\singlespacing
-\setlength{\parskip}{6pt}
-\setlength{\parindent}{0.5cm}
 
 \begin{abstract}
-The abstract should be written after the rest of the paper is completed.
+Multi-agent reinforcement learning has received a lot of updates in cooperative games recently.
+However, research in general non-cooperative games is currently lagging behind.
+Value-based algorithms such as Q-learning have been one of the most widely studied methods in reinforcement learning, and their independent variations for multi-agent environments are popular approaches due to their simplicity and good scalability.
+Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios.
+In this paper, we discuss the feasibility of these algorithms in non-cooperative settings.
+As a result, we explain the conditions a game must satisfy for the algorithms to work.
+We further propose the game Food Chain.
+The game simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis.
+Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games.
 \end{abstract}
 
-\begin{multicols}{2}
-\setlength{\columnsep}{0.5cm}
+\begin{keywords}
+Multi-agent; reinforcement learning; independent value-based learning; game theory; non-cooperative games; competitive games; Nash equilibrium
+\end{keywords}
 
 \section{Introduction}
-Many real-world problems can be modeled as non-cooperative games, such as business competition, political campaigns, stock trading, bidding in auctions, cyber security, and traffic routing.
-These problems feature the competitive nature of games because they share one thing in common, parties with conflicting interests.
-Plenty of research has been done on multi-agent reinforcement learning (MARL) in fully cooperative games.
-There is currently a gap in the non-cooperative setting \cite{zhu2024survey}.
-Most current SOTA MARL algorithms are designed for cooperative games.
-
-In fully cooperative games, there is no conflict between agents.
-If agents can communicate, they are motivated to form a firm alliance that essentially works as a collective mind that is the one and only conscious being.
-Thus, in a game theory sense, a fully cooperative game where agents share information is equivalent to a single agent game where the collective agent tries to find the optimal joint action, given the joint observation.
-To do that in a traditional single agent setting where the environment is stationary and non-adversarial, one only needs to learn the environment mechanics (the state transition function) and the reward function so that the Q-values of an action in a given state can be computed or predicted by a function approximator.
-A Pareto optimal pure (deterministic) strategy is guaranteed to exist in fully cooperative finite games, meaning that the agents can always deterministically choose the optimal action that has the highest payoff.
-The only thing that keeps the agents from forming a cluster in practice is the hardware incapability (computational power, real-world constraints, etc.).
+Recent multi-agent reinforcement learning (MARL) research has mostly focused on fully cooperative games \cite{yu2020benchmarking, papoudakis2020benchmarking, zhu2024survey} where there is no conflict between agents as they share a common reward.
+If we are allowed to share information among the agents, a fully cooperative game reduces to a single-agent game where the collective agent tries to find the optimal joint action, given the joint observation, although the problem is still challenging in practice because of the computational complexity of the joint action brings.
+In a traditional single-agent setting, the agent needs to learn which action gives the best reward.
+A Pareto-optimal pure strategy is guaranteed to exist in fully cooperative finite games, meaning that the agents can always deterministically choose the optimal action that has the highest payoff.
 
@@
 Agents may belong to different parties that would like to have private control to protect their own interests.
-For example, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss, if we omit the long-term economic growth.
+Rational self-interested agents will have to fight for what is best for them, as increasing one agent's reward might lead to the decline of other's gains.
+We can no longer intuitively define the solution for a game as the optimal strategy resulting in the highest reward.
+One of the possible solution concepts is the Nash Equilibrium (NE).
+NEs describe the stable points in games, which are associated with the convergence to optimal behaviour of agents.
 
-For cooperative games, we can always say that 5 rewards is better than 2 rewards, but what about the rewards $\langle 1,5 \rangle$ and  $\langle 10,1 \rangle$ where each agent gets one value in the vector?
-Rational self-interested agents will have to fight for what is best for them.
-Many solutions in game theory are possible.
-One of the commonly accepted solutions for is the Nash equilibrium (NE).
-NEs are normally what we want the agents to converge to because convergence means stability and strict NEs are the only stable points in the game.
-Generally, there may only be a mixed NE, meaning that players adopt randomized strategies consisting of probability distributions over the action space.
+Many real-world problems can be modelled as non-cooperative games, such as business competition, political campaigns, stock trading, bidding in auctions, cyber security, and traffic routing.
+For example, if we omit the long-term economic growth, the stock market is often considered a 0-sum fully competitive game where one person's profit is another person's loss.
+Another example is cyber security, which must study the conflicts between attackers and defenders.
+These problems feature the competitive nature of games because they share one thing in common, parties with conflicting interests.
+Substantial research has been done on MARL in fully cooperative games, and non-cooperative games have received less attention \cite{zhu2024survey, li2024fightladder}.
 
-Currently, most SOTA MARL algorithms cannot learn a mixed policy.
-Largely due to the studied games are mostly fully cooperative or asynchronous.
-As we will show in the experiments, those algorithms will fail to find the optimal action in games specifically designed to challenge the agents' competitive thinking.
-We need to reconsider our approach if we want to build agents that are capable of dealing with non-fully-cooperative environments.
+There are many attempts to conquer games with MARL.
+One of which is independent value-based learning (IVBL).
+IVBL is a family of algorithms that models the multi-agent environments as single-agent environments, and they compute for each agent the expected reward of each action under each state.
+IVBL is simple and highly scalable because it is decentralized.
+It has been adopted for both cooperative \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative} and non-cooperative tasks \cite{bjornsson2009cadiaplayer, jiang2018q, qu2020distributed, kopacz2023evaluating}.
+This paper analyses the performance of IVBL in non-cooperative finite games.
+The analysis considers characteristics of the action space and the NEs of the game, combining both formal analysis and experimental evaluation in selected two-player non-cooperative games.
+The results show NEs with uniformly random policies is a necessary condition for stability convergence of IVBL algorithms in non-cooperative environments.
+These condition highly restricts the wider applicability of IVBL.
 
-\subsection{Aim}
-The aim of this study is to point out that current SOTA algorithms can be misleadingly performing well in non-fully-cooperative environments by both formal analysis and showing specific game scenarios that are designed to challenge the agents from a game theory perspective.
+\subsection{Main contributions}
+\begin{itemize}
+    \item Formal analysis identifying necessary conditions for stability convergence of IVBL which can drive the choice of algorithms in non-cooperative finite games.
+    \item Introducing a non-cooperative game that has the characteristics which challenge the agents from a game theory perspective aligned with our formal analysis.
+    \item Experimental evaluation of IVBL algorithms in selected two-player scenarios in the game proposed above.
+\end{itemize}
+
+\subsection{Contents}
+The remainder of the paper is organized as follows:
+In the next section, we provide the literature review on the key features of non-cooperative games and the most popular MARL algorithms that is related to our work.
+We then state the formal definitions of games and other concepts discussed in this paper in section 3.
+Section 4 analyses the stability of IVBL in non-cooperative games.
+To verify our analysis with experimental results, a new game called Food Chain is introduced in section 5.
+The game will have the key features discussed in our analysis that bring specific challenges to MARL agents.
+Then, we present the design as well as the results of our experiments in section 6.
+Finally, draw our conclusion in the last section.
+
+\section{Tackling Non-Cooperative Games}
+This literature review focuses on key features of non-cooperative games, as well as the MARL approaches to solve these games.
+
+\subsection{Non-cooperative Games}
+In single-agent or fully cooperative environments, there is one reward function for all agents.
+Thus, the agents only need to learn which (joint) action has the highest payoff.
+In non-cooperative games, there may be no optimal joint action that satisfy all agents.
+Many solution concepts to non-cooperative games are discussed in game theory \cite{albrecht2024multi}, such as Pareto optimum, correlated equilibrium, and Nash equilibrium.
+John Nash is famous for his study on non-cooperative games and his concept of Nash Equilibrium (NE) \cite{nash2024non}.
+NE describes a stable situation where no player can benefit from deviating from their current strategy independently.
+NEs offer a framework to characterise and analyse stability in games.
+Stability is associated to the optimal outcome in non-cooperative games and hence the objective agents aim for.
+Generally, there may only be a mixed NE, meaning that players must adopt randomized action-selection policy consisting of probability distributions over the action space.
+This is the mixed policy problem.
+
+Many fully competitive games are studied in the MARL community such as Pommerman \cite{resnick2018pommerman}, Go \cite{silver2017mastering}, StarCraft II \cite{vinyals2019grandmaster}, and Dota 2 \cite{berner2019dota}.
+There are also some general non-cooperative environments such as hide-and-seek \cite{baker2019emergent}, Neural MMO \cite{suarez2021neural}.
+PettingZoo and Melting Pot \cite{terry2021pettingzoo, leibo2021scalable} are two integrated MARL environments including many cooperative and non-cooperative games.
+\cite{kopacz2023evaluating} studied the non-cooperative game predator-and-prey in the PettingZoo library, which is closely related to the game defined in our work.
+However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games.
+In cooperative games, the reward will simply increase as the agents get better at the game.
+This is no longer the case in non-cooperative games where increasing one agent's outcome often results in the drop of another agent's return.
+FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.
+
+\subsection{MARL Algorithms}
+Independent Learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment.
+The agents are essentially unaware of the existence of each other.
+IL is popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}.
+Independent value-based learning (IVBL) algorithms, such as the classic independent Q-Learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling.
+IVBL algorithms including IQL are often considered a baseline and a natural choice in a wide range of MARL studies because of their simplicity and scalability.
+For example, they were studied in various cooperative environments \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative}.
+The problem is that IVBL agents only give us the average reward of each action for each agent, but do not tell us what the action-selection policy look like.
+This is not a problem in single-agent or cooperative environments as we can simply pick the action with the highest return.
+Some studies addressed this issue in non-cooperative games by simply choosing the action with the highest empirical return \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating} and observed reasonably good outcome.
+\cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey.
+Both the algorithm and the game are closely related to our work.
+Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function (\ref{eq:softmax}) to calculate a policy based on the payoff.
+Softmax takes as input a vector and computes a probability distribution over the original vector elements, such that the higher the original value, the higher the probability.
+\begin{equation}
+    \textbf{softmax}([x_1, ..., x_n]) = [\frac{e^{x_1}}{\sum_{i}{e^{x_i}}}, ..., \frac{e^{x_n}}{\sum_{i}{e^{x_i}}}]
+    \label{eq:softmax}
+\end{equation}
+The latter approach intuitively makes sense, as it emphasis the action with higher empirical return.
+However, these approaches have a huge problem that guarantees failure of learning in a wide range of games.
+We will take a closer look in the analysis section.
+
+Unlike IL, joint action learning (JAL) algorithms allow the agent to gain access to the information processed by all other agents, such as their choices of actions, so agents can learn the expected return of joint actions.
+Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games.
+It computes local minimax solutions for each state based on the learned joint Q-values.
+Nash Q-learning \cite{hu2003nash} is a similar algorithm that computes the local NE instead.
+It converges to a NE under highly strict assumptions that reduce the applicability of the approach to a limited set of games.
+Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees.
+\cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games.
+Centralized learning requires full knowledge of the action spaces of all agents, which may not always be available if we are dealing with real-world problems.
+Another major disadvantage is that the number of joint actions is the number of possible combinations of the individual actions.
+This introduces combinatorics to the computational complexity of our algorithms, making them unscalable.
+
+In contrast to the value-based approach, policy-based algorithms, such as the policy gradient methods, work around this issue by learning the policies directly, without putting an estimation on the individual action return.
+Policy gradient methods optimize a parametrized policy function using gradient decent to maximize the return.
+An advantage of policy gradient methods is that they naturally work on continuous action spaces.
+The actor-critic algorithms are a family of RL algorithm which trains a parametrized policy called the actor and a state-value function called the critic to estimate the policy gradient.
+Soft actor-critic (SAC) \cite{haarnoja2018soft} has become one of the most popular algorithms recently.
+Its multi-agent variant (MASAC) has been considered the state-of-the-art (SOTA) baseline \cite{yu2020benchmarking, zhang2020lyapunov}.
+There is also naturally an independent version (ISAC) for MARL.
+The proximal policy optimization algorithm (PPO) \cite{schulman2017proximal} was based on actor-critic as well and got extended to multi-agent environments by some researchers \cite{yu2020benchmarking, yu2022surprising} (MAPPO).
+Surprisingly, the IL version of PPO (IPPO) has proven to substantially outperform MAPPO and other algorithms in some cooperative tasks in StarCraft II \cite{de2020independent}.
+MADDPG \cite{lowe2017multi} is another policy gradient algorithm that is often considered SOTA.
+Because it only works for continuous action spaces, we do not discuss them further in the context of finite games.
+
+Other notable MARL algorithms that have shown good performance include VDN \cite{sunehag2017value} and its improved version QMIX \cite{rashid2020monotonic}.
+The weakness is that they can only work for fully cooperative games where agents share a common reward.
+
+\subsection{Summary}
+As shown in this review, the issue of deriving the policy from the Q-values learned by IVBL agents in non-cooperative games has not been fully addressed by the MARL research community, \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed} made progress in this direction, but they did not study the behaviour of the agents and identify the flaws of the algorithms from a game theory point of view.
+
+\section{MARL Background}
+In this section, we provide preliminaries including the formal definition of a game and NE.
 
 \subsection{Stochastic Game}
-
-A stochastic game is formally defined by a tuple:
-
+A game is formally defined by a tuple:
 $G = \langle I,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
 \begin{itemize}
-    \item $I$: the set of agents, from 1 to \(n\).
+    \item $N = \{1, ..., n\}$: the set of agents, numbered from 1 to $n$.
     \item $S$: the game’s state space.
-    \item $S_0$: the set of possible initial states.
-    \item $S_t$: the set of terminal states.
+    \item $S_0 \subset S$: the set of possible initial states.
+    \item $S_t \subset S$: the set of terminal states.
     \item $\{A_1,\ldots,A_n\}$: one action space for each agent.
-
 Let $a_i \in A_i$ be an action of agent $i$, and $a^n=\langle a_1,\ldots,a_n \rangle \in A^n$ denote a joint action in the joint action space.
-    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent.
-
-    $R_i:S \times A^n \times S \rightarrow \mathbb{R}$
+    \item $\{R_1,\ldots,R_n\}$: one reward function for each agent. $R_i:S \times A^n \times S \rightarrow \mathbb{R}$.
     \item $T$: the state transition function.
-
-    $T:S \times A^n \times S \rightarrow [0,1]$
-
-    It defines the probability of the game transitions from one state to another given the joint action. It must satisfy the axiom of probability:
+$T:S \times A^n \times S \rightarrow [0,1]$.
+It defines the probability of the game transitions from one state to another given the joint action.
+It must satisfy the axiom of probability:
     \begin{equation}
         \forall s \in S,a^n \in A^n: \sum_{s' \in S}{T(s,a^n,s')}=1
-    \label{eq:statetransition}
+        \label{eq:state_transition}
     \end{equation}
-    It is convenient to define $T^*:S \times A^n \rightarrow \Pi_S$, where $\Pi_S$ denotes the set of all probability distribution over the set S.
-
-    Some games incorporate partial observability and define an observations space, which is irrelevant to the purpose of this study.
 \end{itemize}
-
- Starting from a randomly chosen $s_0 \in S_0$, at each timestep $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$.
+Starting from a randomly chosen $s_0 \in S_0$, at each time step $t \in \mathbb{N}$ with the current state $s_t$, each agent observes the current state $s_t$ and chooses an action forming a joint action $a^n=\langle a_1,\ldots,a_n \rangle$.
 The game then moves into the next state $s_{t+1}$, randomly sampled from the distribution P and a reward $R_i$ is assigned to each agent.
 If $s_t \in S_t$, the game ends.
-Otherwise, the game continues in the next timestep.
+Otherwise, the game continues in the next time step.
 
-The goal is to find the optimal strategy (policy) for an agent to pick actions given the current state information they receive from the environment, such that their expected cumulative reward is maximized.
+The way a player plays the game defines a policy (strategy) as a function that computes the probability of selecting each action given the current state, $\pi:S \times A^n \rightarrow [0,1]$.
+A strategy is said to be pure if it is deterministic ($\pi:S \times A^n \rightarrow \{0,1\}$).
+Otherwise, we call it a mixed strategy.
+A set of strategies, one for each player, is called a profile $\pi^n$.
+Given the profile, we can compute the expected reward for each player in the game.
+The goal is to find the optimal strategy for an agent, or even the optimal profile, such that expected cumulative rewards are maximized.
+Cooperative games are special cases where agents share the same reward.
+The optimal profile is the one that maximizes the common reward.
+Because they share the same reward function, we can view the problem as a joint player selecting a joint action and receive a reward, which makes cooperative games essentially single-agent games from a game theory perspective.
+In general non-cooperative games, however, it may not be possible to make everyone happy as their are conflicting interests.
+A more profound definition of optimum need to be discussed in the non-cooperative games section.
 
-To demonstrate the problem with competitive games, let's analyze some simple normal form games.
-A normal form game has only two states, the starting state and the end state.
+Formally, let $R_i^*(\pi^n)$ be the function to compute the expected reward of the i'th player given the profile.
+Let $\pi_i$ be a policy of the i'th player and $\Pi_i$ be the policy space of the i'th player.
+Let $\pi^{-i}$ denote $\pi^n - \{\pi_i\}$, the set of strategies of all other players.
+A profile $\pi^n$ is a NE if
+\begin{equation}
+    \forall i \in N, \forall \pi_i \in \Pi_i: R_i^*(\{\pi_i\} \cup \pi^{-i}) \leqslant R_i^*(\pi^n)
+    \label{eq:nash}
+\end{equation}
+A NE is called strict if $\leqslant$ is replaced by $<$. The idea is that NEs are the stable points in the game, because if an agent can benefit from switching strategy on its own, they would do so.
+
+A game is finite if $|N|, |S|$ and all $|A_i|$ are finite, and the game always ends in a finite number of turns.
+John Nash proved that every finite game has at least one NE.
+To solve a game in the context of this paper means to find at least one NE.
+Games that are not finite can be ill-conditioned and have no NE.
+For example, a game where the player who names the greatest number wins has no best strategy.
+For the rest of the paper, we only refer to finite games.
+A normal form game has only two states, the starting state $S_0$ and the end state $S_1$.
 The game ends immediately after all agents pick an action.
 Thus, they can be represented using the reward matrix.
-Consider rock-paper-scissors (simple one-shot matrix game) in Table~\ref{table:rps} for example.
-Each entry represents the reward for the row player and the column player, respectively.
+In the book on game theory originally published in 1944 \cite{von2007theory}, Von Neumann and Morgenstern showed us that finite games can be translated into normal form games.
+If we encapsulate the following states as a subgame, we end up with a reward matrix for the initial state with the rewards being the expected rewards of the subgames.
+We focus on normal form games because they are easier to work with for the formal analysis.
+Since they only have one state $S_0$ for agents to take actions, we simply write $\pi(a)$ instead of $\pi(S_0, a)$ to denote the probability of taking action $a$.
 
-\noindent
-\begin{minipage}{\columnwidth}
-\captionof{table}{Rock Paper Scissors} 
-\label{table:rps}
-\centering
-\begin{tabular}{c|ccc}
+\section{Stability of IVBL in Non-Cooperative Games}
+In this part of the paper, we show the feasibility of IVBL algorithms in non-cooperative games with formal analysis.
+We demonstrate the necessary conditions the game must satisfy in order to apply the IVBL algorithms.
+
+\begin{table}
+    \begin{minipage}{0.5\textwidth}
+    \tbl{Rock-Paper-Scissors Reward Matrix}{
+    \begin{tabular}{l|ccc}
     & r & p & s\\
@@
     \end{tabular}
+    }
+    \label{table:rps_reward_matrix}
     \end{minipage}
-
-The only NE in the game is that two players play each of the three actions with 1/3 chance.
-If any player deviates from the NE, their expected reward does not increase, and the other player can exploit them by playing the action that counters their most frequent action.
-Here is a more complicated game called biased rock paper scissors.
-
-\noindent
-\begin{minipage}{\columnwidth}
-\captionof{table}{Biased Rock Paper Scissors} 
-\label{table:biasedrps}
-\centering
-\begin{tabular}{c|ccc}
+    \begin{minipage}{0.5\textwidth}
+    \tbl{Biased Rock-Paper-Scissors Reward Matrix}{
+        \begin{tabular}{l|ccc}
         & r & p & s\\
@@
         \end{tabular}
+    }
+    \label{table:biased_rps_reward_matrix}
     \end{minipage}
+\end{table}
 
+To demonstrate the problem with competitive games, let's analyse some simple normal form games.
+Consider the simple one-shot matrix game rock-paper-scissors (RPS) in Table~\ref{table:rps_reward_matrix}.
+Each entry represents the reward for the row player and the column player, respectively.
+The only NE in the game is that two players play each of the three actions with 1/3 chance.
+If any player deviates from the NE, their expected reward does not increase, and the other player can exploit them by playing the action that counters their most frequent action.
+Table~\ref{table:biased_rps_reward_matrix} defines a more complicated game called biased rock-paper-scissors.
 The only NE in the game is that both players adopt the following policy.
@@
     \pi=(P(r)=0.0625,P(p)=0.6250,P(s)=0.3125)
-\label{eq:biasedrpspolicy}
+    \label{eq:biased_rps_policy}
 \end{equation}
 
-The RL algorithm needs to learn $\pi$, not the optimal action because there is no optimal action.
-For competitive play after the learning process, the agent should play randomly according to the learned distribution.
+A well-known direct consequence of (\ref{eq:nash}) in game theory is that at any NE $\pi^n$, given the policy set of the other players $\pi^{i-1}$, player $i$ is indifferent from all actions that are chosen with a positive probability.
+This means that the actions are split into two sets, the worthy actions that are chosen with a positive probability and the worthless ones that will never be piked.
+For example, a variation of RPS could incorporate an always-losing action, where the player receives negative reward once executing it, regardless of the opponent’s action.
+Such an action should never be selected by the players.
+Additionally, the expected reward will be the same for every worthy action.
+In other words, the player can choose the worthy actions arbitrarily, given that other players commit to the NE.
+If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions indeed have the same payoff.
+The above fact can be formally stated as the following.
+Assume $\pi^n$ is a NE.
+Let $Q_i(a_j, \pi^{-i})$ be the expected reward of action $a_j \in A_i$ for agent $i$, given the rest of the profile $\pi^{-i}$.
+Let $A^+_i$ denote the set of worthy actions of agent $i$, namely $A^+_i = \{a | a \in A_i \wedge \pi_i(a) > 0 \}$.
+We have:
+\begin{equation}
+    \begin{aligned}
+    \forall a_j, a_k \in A^+_i, Q_i(a_j, \pi^{-i}) = Q_i(a_k, \pi^{-i})
+    \label{eq:nash_indifferent}
+    \end{aligned}
+\end{equation}
 
-\section{MARL in Competitive Games}
-Algorithms that are carried over from the traditional single agent RL domain, such as Q-learning, do not incorporate randomized policy, even though they may behave randomly during the learning process (using an $\epsilon$-greedy exploration vs.
-exploitation strategy, for example).
-In the single agent setting, the agent only needs to learn the expected long-term return (the Q-value) of each action and deterministically pick the highest.
-Thus, they cannot learn $\pi$.
+Suppose a IVBL algorithm can converge to a NE, then all worthy action's Q-values must converge to the same for an agent.
+It does not matter how the algorithm computes a probability distribution over the actions at the end, it cannot make any discriminations among those worthy actions based on equal statistics.
+Such an algorithm will output at best a uniform distribution over the worthy actions at a NE.
+This contradicts with the assumption that they converge to a NE unless the NE is full of uniformly random policies which is not the case in most games.
+We have shown that pure IVBL approaches cannot work through proof by contradiction.
 
-There is a family of algorithms called Joint Action Learning (JAL) that addresses the problem by incorporating the Q-values for the joint action.
-Minimax Q-learning \cite{littman1994markov} is a JAL algorithm that can be applied to two-player zero-sum games.
-It computes local minimax solutions for each state based on the learned joint Q-values.
-Nash Q-learning \cite{hu2003nash} is a similar algorithm that computes the local NE instead.
-It converges to a NE under highly strict assumptions that is found in almost no games.
-Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees.
-\cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games.
+This gives us a necessary condition for IVBL algorithms to work.
+IVBL algorithms can converge to a NE only if the NE consists of uniformly random policies over the set of all worthy actions.
+Note that this also includes any pure (deterministic) policies.
+It can be formally stated as follows.
+\begin{equation}
+    \exists \ \textbf{a NE} \ \pi^n, \forall \pi_i \in \pi^n, \forall a_j, a_k \in A^+_i, \pi_i(a_j) = \pi_i(a_k)
+    \label{eq:necessary_condition}
+\end{equation}
+Because in cooperative games, there is no need to use mixed strategies, the NEs are essentially pure.
+This means all cooperative games satisfy (\ref{eq:necessary_condition}), which explains why IVBL works in cooperative games.
 
-Some other research \cite{qu2020distributed} tried to address this problem using a manually calculated distribution over the actions based on the payoff, such that the higher the payoff of an action, the more likely the action will be chosen.
-This approach intuitively makes sense.
-The problem is that in any NE, for any action that are chosen with a positive probability, the payoff of the actions are the same.
-If we calculate the expected reward of each action in the biased rock-paper-scissors example where the opponent uses the NE strategy, we will see all three actions have the same payoff.
+Then how did researchers still find the appealing outcome of IVBL in some non-cooperative games \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed}?
+The behaviour of IVBL can be highly problem-dependent, so we provide only some high-level insight here.
+MARL problems studied in the real world can have large state/action spaces.
+For example, a $19 \times 19$ Go board can have roughly $3^{19 \times 19} = 10^{172}$ states.
+It may be too difficult for the agents to converge to a global NE because of the limited computational power.
+Because of the complexity of the environment mechanics itself, the competitiveness of the game is not going to play an enormous role before agents can reliably predict the state transitions.
+Picking the action with the best Q-value still gives reasonably good results, since at least the agents learned to interact with the environment.
 
-Suppose the algorithm can converge to a NE, then all Q-values should be the same (ignoring worthless actions).
-Such an algorithm will output a uniform distribution.
-The algorithm is essentially learning a pure random policy.
-This contradicts with the assumption that they converge to a NE.
-We have shown that approach cannot work through a proof by contradiction.
+\section{The Food Chain Game}
+To verify our formal analysis, we test the algorithms in a game that can turn on and off the necessary condition (\ref{eq:necessary_condition}) described in the previous section.
+We define the mechanics of the game Food Chain.
+An agent in the game represents an animal of some type.
+They can move around the map and feed themselves while trying to avoid enemies.
+Example game scenarios are shown in Figure~\ref{fig:biased_rps_in_food_chain} and Figure~\ref{fig:tree_maze_chase}.
+The game simulates an ecosystem to a certain extend, by defining a food web structure.
+A food web is a cluster of food chains showing what-eats-what in the ecosystem as illustrated in Figure~\ref{fig:food_web}.
+The food web in the game is completely customizable and we can define an arbitrary number of species (agent types) and any number of agents of each type.
+This game is essentially a generalized version of the classic predator-and-prey problem.
+Recent research has tested IQL in the predator-and-prey game and found the algorithm to be seemingly performing well \cite{kopacz2023evaluating}.
+As we stated, these results could be misleading.
 
-Most MARL problems studied are much more than single state matrix games and have astronomically large state/action spaces.
-It is impossible for the agents to converge to a global NE in the first place because of the limited computational power.
-Because of the complexity of the environment that many studies carry out in, picking the action with the best Q-value still gives reasonably good results since at least the agents learned to interact with the environment.
-The competitiveness of the game is not going to play a huge role before the agents can really predict the state transitions.
-Currently, many fully competitive games are studied in the MARL community such as Pommerman \cite{resnick2018pommerman}, Go \cite{silver2017mastering}, StarCraft II \cite{vinyals2019grandmaster}, and Dota 2 \cite{berner2019dota}.
-There are also some general environments that are non-cooperative such as hide-and-seek \cite{baker2019emergent}, Neural MMO \cite{suarez2021neural}, PettingZoo \cite{terry2021pettingzoo}, and Melting Pot \cite{leibo2021scalable}.
-However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games.
-FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.
+\begin{figure}
+\centering
+\includegraphics[width=0.7\textwidth]{images/food_web.png}
+\caption{A simple ecosystem is represented by many food chains showing what-eats-what}
+\label{fig:food_web}
+\end{figure}
 
-\section{Methods}
-We define the mechanics of the game Food Chain.
-I also will have to migrate the tables and charts from my previous Word documents to explain the unique characteristics that make it challenging which are not found in other games (especially compare to the hide and seek game).
+The game is played on a $h \times w$ grid.
+Each agent is placed in one of the tiles.
+There are also obstacles placed on the map.
+These could be randomly generated using a procedural algorithm or predefined.
+The state space $S$ contains all possible configurations of the $h \times w$ game board.
+A game board configuration include the locations of all the agents and other game elements, as well as the current timestep $t$.
+Any state could be an initial state.
+A state is terminal if $t$ reaches some preset maximum timestep.
+The action space $A = \{\textbf{up, down, left, right, pass}\}$ is the same for all agents.
+Agents execute these actions to move along the direction by one cell.
+Agents are not allowed to move into obstacles or exit the map.
+Illegal moves will be regarded as pass.
+If multiple agents try to occupy the same cell, only a randomly selected agent will succeed and the rest will execute the pass action.
+If an agent is moving away from its current cell, then it is allowed for another agent to enter this cell without conflicts, unless two agents are trying to swap location (this rule prevents agents from passing through each other).
+After all agents have executed their actions, if a predator is adjacent to (i.e.
+within a Manhattan distance of 1) a prey, the prey dies and is removed from the game.
+Agents receive some negative reward (-1 by default) when they die.
+Each empty tile also has a chance to grow food (vegetation) each turn.
+Food is part of the food web and can be consumed by some agents when they enter the tile with the food (this is different from agents eating agents where they only need to be adjacent).
+Agents get reward (0.1 by default) when they eat.
+If multiple predators capture a prey at the same time, they must share the reward equally.
+After all the process described above, the game moves to the next state with the updated board configuration.
 
-The game is complex enough to eliminate any tabular approaches.
-The state space is roughly 1060 for just a 9x9 board and can grow exponentially.
-Function approximation methods, such as deep neural networks or Monte Carlo methos must be used to solve the whole game.
-The spacial information on the game board can also be a good challenge to algorithms using CNNs.
+The game is complex enough to eliminate any tabular approaches, a brute-force approach that simply records the value of all possible state-action pairs in a table.
+The size of the state space grows exponentially with respect to the number of agents, the number of species, and the size of the map.
+Function approximation methods, such as deep neural networks or Monte Carlo methods must be used to solve the whole game \cite{silver2016mastering}.
 The game is still simple enough to be human playable.
@@
 Because of the large state space, it is impossible for us to check whether the agents converged to a global NE since it is NP-Hard.
-Yet it has carefully designed mechanics (see my Casual Monthly Progress Report for Nov 2024) which very often force the agents to adopt random policies (and addressing the coordination problem, too).
-These are the leaf nodes of the game tree, which is equivalent to a simple matrix game.
-We can actually check whether the agents converge to NEs in those sub-cases easily.
-So, in summary.
+Yet it has carefully designed mechanics which very often force the agents to address the mixed policy problem.
+We can actually check whether the agents converge to NEs in those sub-cases easily as we will see in the methods section.
+In summary:
 \begin{itemize}
-    \item It is simple enough to let the agents focus on interactions instead of the environment.
-    \item It is complex enough to be interesting.
+    \item It is simple enough to let the agents focus on interactions instead of the environment, yet complex enough to be interesting as it blocks out brute-force algorithms.
+    \item The game is extendable in many dimensions, the number of agents of each type, the number of species, the relationship between the species, and the size of the map.
     \item It forces the agents to constantly address the mixed strategy problem.
-    \item We can analyze the performance of the agents with game theory.
+    \item We can analyse the performance of the agents with game theory.
 \end{itemize}
 
-\section{Results}
-Figure \ref{fig:hopperreward}, \ref{fig:frogreward}, \ref{fig:snakereward} shows the average test return for each type of agent after training in a randomly generated game environment.
-Figure \ref{fig:scenhopperreward}, \ref{fig:scenfrogreward} demonstrates that in the carefully set up matching pennies scenario.
-The results show that the algorithms could not reach the optimal reward which can be computed using linear algebra (which is also another strong aspect of the game.
-we can analyze the game formally and see whether the agents reached optimality).
-More formal analysis coming...
+\section{Experimental Evaluation}
+Too see how Food Chain can turn on/off the necessary condition (\ref{eq:necessary_condition}) for IVBL to converge, we design two scenarios.
+The key difference between the two is that one satisfies the condition (\ref{eq:necessary_condition}) and the other does not.
+We define a simple food web consisting frogs and hoppers.
+Hoppers (represented by h) can consume vegetation (drawn as *) and will be eaten by frogs (f).
+Both scenarios will turn off random food generation.
 
-\noindent
-\begin{minipage}{\columnwidth}
+\begin{figure}
 \centering
-\includesvg[width=1\textwidth]{plots/rewards_global_hopper.svg}
-\captionof{figure}{hopper rewards}
-\label{fig:hopperreward}
+    \begin{minipage}{0.48\textwidth}
+        \centering
+        \includegraphics[width=0.93\textwidth]{images/biased_rps.png}
+        \caption{Biased RPS in Food Chain. The frog (f) is trying to capture the hopper (h). The only NE in the game contains non-uniformly random mixed policies.}
+        \label{fig:biased_rps_in_food_chain}
+    \end{minipage}\hfill
+    \begin{minipage}{0.48\textwidth}
+        \centering
+        \includegraphics[width=0.9\textwidth]{images/tree_maze_chase.png}
+        \caption{Tree Maze Chase in Food Chain. The frog (f) is chasing the hooper (h) in the maze. There is at least one optimal pure policy for both players.}
+        \label{fig:tree_maze_chase}
     \end{minipage}
+\end{figure}
 
-\noindent
-\begin{minipage}{\columnwidth}
+\subsection{Biased Rock-Paper-Scissors in Food Chain}
+This scenario forces the agents to address the mixed strategy problem.
+We set up the environment as shown in Figure~\ref{fig:biased_rps_in_food_chain}.
+Since all agents have the same mobility, a single predator can only capture a prey by trapping them at a corner.
+All predators, unless working in a team, have to face this subcase at some point during a general Food Chain game.
+Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} in the appendix shows all possible next states.
+We can see that in some cases the hopper can escape.
+If they both pass, they will face the exact same situation again.
+For simplicity, we set the max timestep to 1, so the game ends after one move.
+The scenario becomes a normal form mini game.
+To break the symmetry so that condition (\ref{eq:necessary_condition}) is violated, we add food in one cell, making the scenario essentially biased RPS.
+We also raise the feeding reward for both agents to 0.5 to highlight the asymmetry.
+Let hopper be the row player.
+The reward matrix is shown in Table~\ref{table:biased_rps_in_food_chain_reward_matrix} in the appendix.
+
+We can verify that there is a unique NE in the game.
+\begin{equation}
+    \pi_h=(P(r)=\frac{1}{3},P(d)=\frac{1}{3},P(p)=\frac{1}{3})
+    \label{eq:biased_rps_in_food_chain_hopper_policy}
+\end{equation}
+\begin{equation}
+    \pi_f=(P(l)=\frac{1}{13},P(u)=\frac{6}{13},P(p)=\frac{6}{13})
+    \label{eq:biased_rps_in_food_chain_frog_policy}
+\end{equation}
+The expected reward for the hooper and the frog should be $-7/13 \approx -0.538$ and $1/3 \approx 0.333$, respectively.
+
+\subsection{Tree Maze Chase in Food Chain}
+This scenario essentially turns off the mixed strategy problem, so the condition (\ref{eq:necessary_condition}) is met. We set up the game environment as follows.
+
+This map setup resembles a tree structure (no cycles).
+It's easy to see that there is a pure NE in the game, because there is always an optimal action for both players.
+The agents don't need to adopt mixed policies.
+The hooper should go for the tunnel with more food and the frog should simply chase the hopper.
+This means the necessary condition (\ref{eq:necessary_condition}) for IVBL to converge is met.
+
+\subsection{Results and Discussion}
+In both scenarios, we test the IVBL algorithm IQL and the independent policy-based algorithm IPPO.
+The algorithm collects 640 timesteps and train the neural network in each iteration.
+After each training iteration, 50 evaluation games are run during the evaluation phase.
+The plots show the average reward for each agent during the evaluation phase.
+Each experiment is run 20 times to mitigate the effect of randomness and the graphs show the boots trapped 95\% confidence interval.
+In total, each data point on the graph is the average result of $20 \times 50 = 1000$ evaluation games.
+
+Figure~\ref{fig:biased_rps_in_food_chain_rewards} demonstrates the performance of the algorithms in the biased RPS scenario where the necessary condition (\ref{eq:necessary_condition}) for IVBL to converge is not met.
+The results show that the IVBL algorithm IQL failed to converge to anything and are struggling to stabilize itself.
+The hopper's reward fluctuates between -1 and 0, and the frog reward oscillates in the range [0.1, 0.5].
+These values can be 100\% away from what it should get (-0.538 and 0.333).
+For comparison, the independent policy-based learning algorithm IPPO achieves a reward that stays more in the optimal range and is far more stable which indicates the agents are converging towards the NE.
+
+In contrast, IVBL does significantly better in the tree maze scenario, as shown in Figure~\ref{fig:tree_maze_chase_rewards}.
+This time, the game scenario satisfies the condition (\ref{eq:necessary_condition}).
+Although still unstable, IQL can roughly reach the optimal reward at the end.
+The optimal reward for the hopper and the frog are -0.8 and 0.1, respectively.
+Even in this case, the policy-based method still appears to be more reliable than IVBL.
+
+\begin{figure}
 \centering
-\includesvg[width=1\textwidth]{plots/rewards_global_frog.svg}
-\captionof{figure}{frog rewards}
-\label{fig:frogreward}
-\end{minipage}
+\subfloat[Biased RPS Hopper Rewards]{
+    \resizebox*{0.45\textwidth}{!}{
+        \includesvg{plots/food_chain_biased_rps_hopper_return.svg}
+    }
+}
+\subfloat[Biased RPS Frog Rewards]{
+    \resizebox*{0.45\textwidth}{!}{
+        \includesvg{plots/food_chain_biased_rps_frog_return.svg}
+    }
+}
+\caption{Biased RPS Rewards}
+\label{fig:biased_rps_in_food_chain_rewards}
+\end{figure}
 
-\noindent
-\begin{minipage}{\columnwidth}
+\begin{figure}
 \centering
-\includesvg[width=1\textwidth]{plots/rewards_global_snake.svg}
-\captionof{figure}{snake rewards}
-\label{fig:snakereward}
-\end{minipage}
-
-\noindent
-\begin{minipage}{\columnwidth}
-\centering
-\includesvg[width=1\textwidth]{plots/rewards_scen_hopper.svg}
-\captionof{figure}{snake rewards}
-\label{fig:scenhopperreward}
-\end{minipage}
-
-\noindent
-\begin{minipage}{\columnwidth}
-\centering
-\includesvg[width=1\textwidth]{plots/rewards_scen_frog.svg}
-\captionof{figure}{snake rewards}
-\label{fig:scenfrogreward}
-\end{minipage}
-
-\section{Discussion}
+\subfloat[Biased RPS Hopper Rewards]{
+    \resizebox*{0.45\textwidth}{!}{
+        \includesvg{plots/food_chain_tree_maze_chase_hopper_return.svg}
+    }
+}
+\subfloat[Biased RPS Frog Rewards]{
+    \resizebox*{0.45\textwidth}{!}{
+        \includesvg{plots/food_chain_tree_maze_chase_frog_return.svg}
+    }
+}
+\caption{Tree Maze Chase Rewards}
+\label{fig:tree_maze_chase_rewards}
+\end{figure}
 
 \section{Conclusion}
+We have identified a necessary condition (\ref{eq:necessary_condition}) for IVBL algorithms to converge to a NE in finite games.
+We further proposed the game Food Chain which has certain traits that can verify our analysis.
+The results show that IVBL algorithms perform poorly in scenarios where the condition (\ref{eq:necessary_condition}) is violated.
+
+\section*{Acknowledgements}
+We thank the creators of the MARL library BenchMARL \cite{bettini2024benchmarl}, which was used to run the experiments.
 
@@
 
-% Prints bibliography
-\printbibliography
+\bibliographystyle{tfnlm}
+\bibliography{ref}
 
-\end{multicols}
+\pagebreak
+\appendix
+
+\section{Biased RPS in Food Chain}
+Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} shows all the possible next state of the Biased RPS scenario in Food Chain.
+Table~\ref{table:biased_rps_in_food_chain_reward_matrix} describes the reward matrix of the scenario (note that if both agents try to move into the food tile, the hopper only has a 50\% chance to get food).
+
+\begin{figure}
+\centering
+\includegraphics{images/biased_rps_possible_next_states.png}
+\caption{Possible Next States in Biased RPS in Food Chain (omitting illegal moves)}
+\label{fig:biased_rps_in_food_chain_possible_next_states}
+\end{figure}
+
+\begin{table}
+\tbl{Biased RPS in Food Chain Reward Matrix}{
+    \begin{tabular}{l|ccc}
+        & left & up & pass\\
+        \hline
+        right & 0.5,0 & -0.75,0.5 & -0.5,0.5 \\
+        down & -1,0.5 & 0,0 & -1,0.5 \\
+        pass & -1,0.5 & -1,0.5 & 0,0 \\
+    \end{tabular}
+}
+\label{table:biased_rps_in_food_chain_reward_matrix}
+\end{table}
+
 \end{document}

### 2025-06-05 paper/paper.tex
Update paper.tex

@@
 \usepackage{svg}
-
+%\usepackage{biblatex}
 \usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
@@
 However, research in general non-cooperative games is currently lagging behind.
-Value-based algorithms such as Q-learning have been one of the most widely studied methods in reinforcement learning, and their independent variations for multi-agent environments are popular approaches due to their simplicity and good scalability.
+Independent value-based algorithms such as independent Q-learning have been one of the most widely studied methods in the field due to their simplicity and scalability.
 Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios.
@@
 As a result, we explain the conditions a game must satisfy for the algorithms to work.
-We further propose the game Food Chain.
-The game simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis.
-Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games.
+We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis.
+Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
 \end{abstract}
@@
 A game is formally defined by a tuple:
-$G = \langle I,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
+$G = \langle N,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
 \begin{itemize}

### 2025-06-07 paper/paper.tex
paper v7

@@
 \name{Yang Li\textsuperscript{a}\thanks{CONTACT Yang Li. Email: yang4.li@uwe.ac.uk}, Marco Perez Hernandez\textsuperscript{a} and Mehmet Emin Aydin\textsuperscript{a}}
-\affil{\textsuperscript{a} University of the West of England, Coldharbour Lane, Frenchay Campus, Bristol, UK}
+\affil{\textsuperscript{a} University of the West of England, Frenchay Campus, Coldharbour Lane, Bristol, UK}
 }
@@
 In this paper, we discuss the feasibility of these algorithms in non-cooperative settings.
-As a result, we explain the conditions a game must satisfy for the algorithms to work.
-We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our formal analysis.
+As a result, we explain the conditions that a game must satisfy for the algorithms to work.
+We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our analysis.
 Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
@@
 Recent multi-agent reinforcement learning (MARL) research has mostly focused on fully cooperative games \cite{yu2020benchmarking, papoudakis2020benchmarking, zhu2024survey} where there is no conflict between agents as they share a common reward.
-If we are allowed to share information among the agents, a fully cooperative game reduces to a single-agent game where the collective agent tries to find the optimal joint action, given the joint observation, although the problem is still challenging in practice because of the computational complexity of the joint action brings.
+When agents are allowed to share information, we can apply single-agent reinforcement learning (RL) algorithms to a fully cooperative game where the collective agent tries to find the optimal joint action, given the joint observation, although the problem is still challenging in practice because of what the computational complexity of the joint action brings.
 In a traditional single-agent setting, the agent needs to learn which action gives the best reward.
@@
 Agents may belong to different parties that would like to have private control to protect their own interests.
-Rational self-interested agents will have to fight for what is best for them, as increasing one agent's reward might lead to the decline of other's gains.
+Rational and self-interested agents will have to fight for what is best for them, as increasing one agent's reward might lead to the decline of other's gains.
 We can no longer intuitively define the solution for a game as the optimal strategy resulting in the highest reward.
-One of the possible solution concepts is the Nash Equilibrium (NE).
+One of the possible solution concepts is the Nash equilibrium (NE).
 NEs describe the stable points in games, which are associated with the convergence to optimal behaviour of agents.
@@
 
-There are many attempts to conquer games with MARL.
+Some studies have explored the application of MARL to solve games.
 One of which is independent value-based learning (IVBL).
@@
 
-\subsection{Main contributions}
+%\subsection{Main contributions}
+The main contributions of this work include the following points:
 \begin{itemize}
-    \item Formal analysis identifying necessary conditions for stability convergence of IVBL which can drive the choice of algorithms in non-cooperative finite games.
-    \item Introducing a non-cooperative game that has the characteristics which challenge the agents from a game theory perspective aligned with our formal analysis.
+    \item Behavioural analysis identifying necessary conditions for a stable convergence of IVBL which can drive the choice of algorithms in non-cooperative finite games.
+    \item Introducing a non-cooperative game that has the characteristics which challenge the agents from a game theory perspective aligned with our analysis.
     \item Experimental evaluation of IVBL algorithms in selected two-player scenarios in the game proposed above.
@@
 
-\subsection{Contents}
+%\subsection{Contents}
 The remainder of the paper is organized as follows:
-In the next section, we provide the literature review on the key features of non-cooperative games and the most popular MARL algorithms that is related to our work.
-We then state the formal definitions of games and other concepts discussed in this paper in section 3.
+The next section provides the literature review on the key features of non-cooperative games and the most popular MARL algorithms that is related to our work.
+We then state the formal definitions of games and other concepts discussed in this paper in Section 3.
 Section 4 analyses the stability of IVBL in non-cooperative games.
-To verify our analysis with experimental results, a new game called Food Chain is introduced in section 5.
+To verify our analysis with experimental results, a new game called Food Chain is introduced in Section 5.
 The game will have the key features discussed in our analysis that bring specific challenges to MARL agents.
-Then, we present the design as well as the results of our experiments in section 6.
-Finally, draw our conclusion in the last section.
+Then, we present the design as well as the results of our experiments and relevant discussions in Section 6.
+Finally, the drawn conclusions pulled in the last section.
 
-\section{Tackling Non-Cooperative Games}
+\section{Background and Literature Review}
 This literature review focuses on key features of non-cooperative games, as well as the MARL approaches to solve these games.
 
-\subsection{Non-cooperative Games}
+\subsection{Non-Cooperative Games}
 In single-agent or fully cooperative environments, there is one reward function for all agents.
@@
 In non-cooperative games, there may be no optimal joint action that satisfy all agents.
-Many solution concepts to non-cooperative games are discussed in game theory \cite{albrecht2024multi}, such as Pareto optimum, correlated equilibrium, and Nash equilibrium.
-John Nash is famous for his study on non-cooperative games and his concept of Nash Equilibrium (NE) \cite{nash2024non}.
-NE describes a stable situation where no player can benefit from deviating from their current strategy independently.
-NEs offer a framework to characterise and analyse stability in games.
-Stability is associated to the optimal outcome in non-cooperative games and hence the objective agents aim for.
+Several problem solving concepts/approaches for non-cooperative games are discussed in game theory \cite{albrecht2024multi}.
+These include Pareto optimum, Nash equilibrium, and correlated equilibrium.
+Nash Equilibrium (NE), proposed by John Nash, is a renown approach for non-cooperative games \cite{nash2024non}.
+NE offers a framework to characterise and analyse stability in games and describes a stability condition with which no player can benefit from deviating from their current strategy independently.
+The stability is associated to the optimal outcome in non-cooperative games and hence the objective agents aim for.
 Generally, there may only be a mixed NE, meaning that players must adopt randomized action-selection policy consisting of probability distributions over the action space.
@@
 PettingZoo and Melting Pot \cite{terry2021pettingzoo, leibo2021scalable} are two integrated MARL environments including many cooperative and non-cooperative games.
-\cite{kopacz2023evaluating} studied the non-cooperative game predator-and-prey in the PettingZoo library, which is closely related to the game defined in our work.
+Kopazs et al.
+\cite{kopacz2023evaluating} studied the non-cooperative game predator-and-prey in the PettingZoo library, which is closely related to the game defined in this work.
 However, we lack a standard benchmark for MARL algorithms in non-cooperative games due to the difficulty of measuring agents' performance in general non-cooperative games.
-In cooperative games, the reward will simply increase as the agents get better at the game.
-This is no longer the case in non-cooperative games where increasing one agent's outcome often results in the drop of another agent's return.
+In cooperative games, the reward simply increases as the agents do better at the game.
+However, this is not the case in non-cooperative games where increasing one agent's outcome often results in the drop of another agent's return.
 FightLadder \cite{li2024fightladder} used the Elo rating system to test MARL algorithms but they were restricted to two-player zero-sum fighting games.
@@
 \subsection{MARL Algorithms}
-Independent Learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment.
-The agents are essentially unaware of the existence of each other.
-IL is popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}.
-Independent value-based learning (IVBL) algorithms, such as the classic independent Q-Learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling.
-IVBL algorithms including IQL are often considered a baseline and a natural choice in a wide range of MARL studies because of their simplicity and scalability.
+Independent learning (IL) algorithms that are carried over from the traditional single-agent RL domain treat other agents as part of the environment.
+The agents are essentially unaware of the existence of one another.
+IL algorithms are popular for handling many complex problems \cite{papoudakis2020benchmarking, gupta2017cooperative, de2020independent, palmer2020independent}.
+Among these, independent value-based learning (IVBL) algorithms, such as the classic independent Q-learning (IQL) \cite{tan1993multi}, learn the expected long-term cumulative return (the Q-value) of each action at each state through sampling.
+IVBL algorithms including IQL are often considered as baseline and natural choice in a wide range of MARL studies because of their simplicity and scalability.
 For example, they were studied in various cooperative environments \cite{foerster2017stabilising, omidshafiei2017deep, palmer2017lenient, palmer2018negative}.
-The problem is that IVBL agents only give us the average reward of each action for each agent, but do not tell us what the action-selection policy look like.
+The problem is that IVBL agents only help produce the average reward of each action for each agent, but do not pay attention on what the action-selection policy.
 This is not a problem in single-agent or cooperative environments as we can simply pick the action with the highest return.
 Some studies addressed this issue in non-cooperative games by simply choosing the action with the highest empirical return \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating} and observed reasonably good outcome.
-\cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey.
+Kopacz et al \cite{kopacz2023evaluating} has examined IVBL in the non-cooperative game predator-and-prey.
 Both the algorithm and the game are closely related to our work.
-Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function (\ref{eq:softmax}) to calculate a policy based on the payoff.
-Softmax takes as input a vector and computes a probability distribution over the original vector elements, such that the higher the original value, the higher the probability.
+Other researchers \cite{qu2020distributed} tried to address this problem using the softmax function given in Equation~(\ref{eq:softmax}) to calculate a policy based on the payoff.
+Softmax takes a vector as input and computes a probability distribution over the original vector entries, such that the higher the original value, the higher the probability.
 \begin{equation}
@@
 \end{equation}
-The latter approach intuitively makes sense, as it emphasis the action with higher empirical return.
+The latter approach emphasis the action with higher empirical return.
 However, these approaches have a huge problem that guarantees failure of learning in a wide range of games.
@@
 Correlated Q-learning \cite{greenwald2003correlated} computes a correlated equilibrium and has no known convergence guarantees.
-\cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games.
+Zinkevich et al \cite{zinkevich2005cyclic} showed that JAL cannot converge in certain types of ill-conditioned games.
 Centralized learning requires full knowledge of the action spaces of all agents, which may not always be available if we are dealing with real-world problems.
@@
 There is also naturally an independent version (ISAC) for MARL.
-The proximal policy optimization algorithm (PPO) \cite{schulman2017proximal} was based on actor-critic as well and got extended to multi-agent environments by some researchers \cite{yu2020benchmarking, yu2022surprising} (MAPPO).
+The proximal policy optimization algorithm (PPO) \cite{schulman2017proximal} was based on actor-critic as well and was extended to multi-agent environments by some researchers \cite{yu2020benchmarking, yu2022surprising} (MAPPO).
 Surprisingly, the IL version of PPO (IPPO) has proven to substantially outperform MAPPO and other algorithms in some cooperative tasks in StarCraft II \cite{de2020independent}.
@@
 \subsection{Summary}
-As shown in this review, the issue of deriving the policy from the Q-values learned by IVBL agents in non-cooperative games has not been fully addressed by the MARL research community, \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed} made progress in this direction, but they did not study the behaviour of the agents and identify the flaws of the algorithms from a game theory point of view.
+As shown in this review, the issue of deriving the policy from the Q-values learned by IVBL agents in non-cooperative games has not been fully addressed by the MARL community, \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed} made progress in this direction, but they did not analyse the behaviour of the agents and identify the flaws from a game theory point of view.
 
-\section{MARL Background}
+\section{MARL Preliminaries}
 In this section, we provide preliminaries including the formal definition of a game and NE.
@@
 A game is formally defined by a tuple:
+
 $G = \langle N,S,S_0,S_t,\{A_1,\ldots,A_n\},\{R_1,\ldots,R_n\},T \rangle$
@@
 The optimal profile is the one that maximizes the common reward.
-Because they share the same reward function, we can view the problem as a joint player selecting a joint action and receive a reward, which makes cooperative games essentially single-agent games from a game theory perspective.
-In general non-cooperative games, however, it may not be possible to make everyone happy as their are conflicting interests.
+Because they share the same reward function, the problem can be formulated as a joint player selecting a joint action and receives a reward, which makes cooperative games essentially single-agent games from a game theory perspective.
+In general non-cooperative games, however, it may not be possible to make everyone happy due to the conflicting interests.
 A more profound definition of optimum need to be discussed in the non-cooperative games section.
 
-Formally, let $R_i^*(\pi^n)$ be the function to compute the expected reward of the i'th player given the profile.
-Let $\pi_i$ be a policy of the i'th player and $\Pi_i$ be the policy space of the i'th player.
+Formally, let $R_i^*(\pi^n)$ be the function to compute the expected reward of the $i^{th}$ player given the profile.
+Let $\pi_i$ be a policy of the $i^{th}$ player and $\Pi_i$ be the policy space of the $i^{th}$ player.
 Let $\pi^{-i}$ denote $\pi^n - \{\pi_i\}$, the set of strategies of all other players.
@@
 \end{equation}
-A NE is called strict if $\leqslant$ is replaced by $<$. The idea is that NEs are the stable points in the game, because if an agent can benefit from switching strategy on its own, they would do so.
+The idea is that NEs are the stable points in the game, because if an agent can benefit from switching strategy on its own, they would do so.
 
@@
 Thus, they can be represented using the reward matrix.
-In the book on game theory originally published in 1944 \cite{von2007theory}, Von Neumann and Morgenstern showed us that finite games can be translated into normal form games.
+In the book on game theory originally published in 1944 \cite{von2007theory}, Von Neumann and Morgenstern showed that finite games can be translated into normal form games.
 If we encapsulate the following states as a subgame, we end up with a reward matrix for the initial state with the rewards being the expected rewards of the subgames.
-We focus on normal form games because they are easier to work with for the formal analysis.
+We focus on normal form games because they are easier to work with for the analysis.
 Since they only have one state $S_0$ for agents to take actions, we simply write $\pi(a)$ instead of $\pi(S_0, a)$ to denote the probability of taking action $a$.
@@
 \section{Stability of IVBL in Non-Cooperative Games}
-In this part of the paper, we show the feasibility of IVBL algorithms in non-cooperative games with formal analysis.
+In this part of the paper, we show the feasibility of IVBL algorithms in non-cooperative games with behavioural analysis.
 We demonstrate the necessary conditions the game must satisfy in order to apply the IVBL algorithms.
@@
     \end{minipage}
+    \tabnote{The reward matrix is used to define the reward for each player for each choice of action. The row player chooses a row and gets the first reward. Vice versa for the column player.}
 \end{table}
 
-To demonstrate the problem with competitive games, let's analyse some simple normal form games.
+To demonstrate the problem with competitive games, we study two simple normal form games.
 Consider the simple one-shot matrix game rock-paper-scissors (RPS) in Table~\ref{table:rps_reward_matrix}.
@@
 
-A well-known direct consequence of (\ref{eq:nash}) in game theory is that at any NE $\pi^n$, given the policy set of the other players $\pi^{i-1}$, player $i$ is indifferent from all actions that are chosen with a positive probability.
+A well-known direct consequence of Equation~(\ref{eq:nash}) in game theory is that at any NE $\pi^n$, given the policy set of the other players $\pi^{i-1}$, player $i$ is indifferent from all actions that are chosen with a positive probability \cite{nash2024non}.
 This means that the actions are split into two sets, the worthy actions that are chosen with a positive probability and the worthless ones that will never be piked.
@@
 Such an algorithm will output at best a uniform distribution over the worthy actions at a NE.
-This contradicts with the assumption that they converge to a NE unless the NE is full of uniformly random policies which is not the case in most games.
-We have shown that pure IVBL approaches cannot work through proof by contradiction.
+This contradicts with the assumption that they converge to a NE unless the NE is full of uniformly random policies which is not the case in most of the games.
+We have shown that pure IVBL approaches does not work through proof by contradiction.
 
-This gives us a necessary condition for IVBL algorithms to work.
+This sets a necessary condition for IVBL algorithms to work.
 IVBL algorithms can converge to a NE only if the NE consists of uniformly random policies over the set of all worthy actions.
@@
 It can be formally stated as follows.
+Let $\Pi^{n}$ be the set of all possible profiles, and $\Pi^{n*}$ be the subset containing all the NEs.
 \begin{equation}
-    \exists \ \textbf{a NE} \ \pi^n, \forall \pi_i \in \pi^n, \forall a_j, a_k \in A^+_i, \pi_i(a_j) = \pi_i(a_k)
+    \exists \pi^n \in \Pi^{n*}, \forall \pi_i \in \pi^n, \forall a_j, a_k \in A^+_i, \pi_i(a_j) = \pi_i(a_k)
     \label{eq:necessary_condition}
@@
 Because in cooperative games, there is no need to use mixed strategies, the NEs are essentially pure.
-This means all cooperative games satisfy (\ref{eq:necessary_condition}), which explains why IVBL works in cooperative games.
+This means that all cooperative games satisfy Equation~(\ref{eq:necessary_condition}), which explains why IVBL works in cooperative games.
+Additionally, according to Zermelo's theorem \cite{schwalbe2001zermelo}, perfect-information two-player finite asynchronous (players take turns) games such as chess or Go also satisfy the condition, because they also have a pure NE.
 
-Then how did researchers still find the appealing outcome of IVBL in some non-cooperative games \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed}?
-The behaviour of IVBL can be highly problem-dependent, so we provide only some high-level insight here.
+Then how did the researchers still find the appealing outcome of IVBL in some non-cooperative games \cite{bjornsson2009cadiaplayer, jiang2018q, kopacz2023evaluating, qu2020distributed}?
+The behaviour of IVBL can be highly problem-dependent, so we provide only high-level insights here.
 MARL problems studied in the real world can have large state/action spaces.
@@
 \section{The Food Chain Game}
-To verify our formal analysis, we test the algorithms in a game that can turn on and off the necessary condition (\ref{eq:necessary_condition}) described in the previous section.
+To verify our analysis, we test the algorithms in a game that can turn on and off the necessary condition given Equation~(\ref{eq:necessary_condition}) described in the previous section.
 We define the mechanics of the game Food Chain.
@@
 The game simulates an ecosystem to a certain extend, by defining a food web structure.
-A food web is a cluster of food chains showing what-eats-what in the ecosystem as illustrated in Figure~\ref{fig:food_web}.
+A food web is a cluster of food chains showing what-eats-what in the ecosystem as illustrated in Figure~\ref{fig:food_chain_game}.
 The food web in the game is completely customizable and we can define an arbitrary number of species (agent types) and any number of agents of each type.
@@
 \centering
-\includegraphics[width=0.7\textwidth]{images/food_web.png}
-\caption{A simple ecosystem is represented by many food chains showing what-eats-what}
-\label{fig:food_web}
+\subfloat[A Food Web]{
+    \resizebox*{0.5\textwidth}{!}{
+        \includegraphics{images/food_web.png}
+    }
+}
+\subfloat[The Food Chain Game]{
+    \resizebox*{0.5\textwidth}{!}{
+        \includegraphics{images/food_chain_game.png}
+    }
+}
+\caption{(a), A food web represents a simple ecosystem.
+The food web is a cluster of multiple food chains showing what-eats-what.
+The arrows point in the direction of the energy flow.
+(b), In the food chain game, an agent represents an individual animal of some type.
+They navigate on a playground trying to feed themselves while avoiding enemies.}
+\label{fig:food_chain_game}
 \end{figure}
 
-The game is played on a $h \times w$ grid.
+The game is played on a $w \times h$ grid.
 Each agent is placed in one of the tiles.
@@
 These could be randomly generated using a procedural algorithm or predefined.
-The state space $S$ contains all possible configurations of the $h \times w$ game board.
-A game board configuration include the locations of all the agents and other game elements, as well as the current timestep $t$.
+The state space $S$ contains all possible configurations of the $w \times h$ game board.
+A game board configuration includes the current timestep $t$, as well as the locations of all the agents, vegetation and obstacles.
+All positions $(x, y)$ must be unique and satisfy $1 \le x \le w, 1 \le y \le h$.
 Any state could be an initial state.
@@
 The action space $A = \{\textbf{up, down, left, right, pass}\}$ is the same for all agents.
-Agents execute these actions to move along the direction by one cell.
-Agents are not allowed to move into obstacles or exit the map.
-Illegal moves will be regarded as pass.
+The state transition function maps the current state and the choices of actions to the next state by executing the agents' actions in a natural way.
+Agents choose their actions to move along the direction by one cell.
+Moves resulting in an illegal position will be regarded as pass.
 If multiple agents try to occupy the same cell, only a randomly selected agent will succeed and the rest will execute the pass action.
 If an agent is moving away from its current cell, then it is allowed for another agent to enter this cell without conflicts, unless two agents are trying to swap location (this rule prevents agents from passing through each other).
-After all agents have executed their actions, if a predator is adjacent to (i.e.
+The reward function works as follows.
+After the state transition, if a predator is adjacent to (i.e.
 within a Manhattan distance of 1) a prey, the prey dies and is removed from the game.
 Agents receive some negative reward (-1 by default) when they die.
-Each empty tile also has a chance to grow food (vegetation) each turn.
-Food is part of the food web and can be consumed by some agents when they enter the tile with the food (this is different from agents eating agents where they only need to be adjacent).
+Each empty tile also has a chance to grow vegetation each turn.
+Vegetation is part of the food web and will be destroyed when an agent enter the tile.
+If the vegetation is a prey to the agent, it will be consumed (this is different from agents eating agents where they only need to be adjacent).
 Agents get reward (0.1 by default) when they eat.
 If multiple predators capture a prey at the same time, they must share the reward equally.
-After all the process described above, the game moves to the next state with the updated board configuration.
 
-The game is complex enough to eliminate any tabular approaches, a brute-force approach that simply records the value of all possible state-action pairs in a table.
-The size of the state space grows exponentially with respect to the number of agents, the number of species, and the size of the map.
-Function approximation methods, such as deep neural networks or Monte Carlo methods must be used to solve the whole game \cite{silver2016mastering}.
+The size of the state space grows exponentially with respect to the size of the map and the number of agents.
+The number of possible states in a game with $n$ agents playing on a board of $k$ cells is lower bounded by $C{k \choose n}$.
+This computational complexity eliminates any tabular method, a brute-force approach that simply records the value of all possible state-action pairs in a table.
+Function approximation methods, such as deep neural networks or Monte Carlo methods should be used to solve the whole game \cite{silver2016mastering}.
 The game is still simple enough to be human playable.
-The mechanics of the game (the state transition function) is very straightforward and easy for the agents to learn, so that they can focus on agent interactions.
-Because of the large state space, it is impossible for us to check whether the agents converged to a global NE since it is NP-Hard.
-Yet it has carefully designed mechanics which very often force the agents to address the mixed policy problem.
-We can actually check whether the agents converge to NEs in those sub-cases easily as we will see in the methods section.
+The mechanics of the game (the state transition function) is straightforward, so that agents can focus on interactions.
+Because of the large state space, it is difficult for us to check whether the agents converged to a global NE since it is NP-Hard.
+However, the way predators hunt for preys often force agents to address the mixed policy problem, as we will explain in the next section with two example scenarios illustrated in Figure~\ref{fig:biased_rps_in_food_chain} and Figure~\ref{fig:tree_maze_chase}.
+We can actually check whether the agents converge to NEs in those subcases.
 In summary:
 \begin{itemize}
-    \item It is simple enough to let the agents focus on interactions instead of the environment, yet complex enough to be interesting as it blocks out brute-force algorithms.
+    \item The navigation in the environment where agents are placed is simple enough, so the agents can focus on learning interactions instead of the state transitions, yet the game is still complex enough to be interesting as it blocks out brute-force algorithms.
     \item The game is extendable in many dimensions, the number of agents of each type, the number of species, the relationship between the species, and the size of the map.
-    \item It forces the agents to constantly address the mixed strategy problem.
+    \item The game forces the agents to constantly address the mixed strategy problem.
     \item We can analyse the performance of the agents with game theory.
@@
 The key difference between the two is that one satisfies the condition (\ref{eq:necessary_condition}) and the other does not.
-We define a simple food web consisting frogs and hoppers.
+We define a simple food web consisting of frogs and hoppers.
 Hoppers (represented by h) can consume vegetation (drawn as *) and will be eaten by frogs (f).
@@
         \centering
-        \includegraphics[width=0.93\textwidth]{images/biased_rps.png}
-        \caption{Biased RPS in Food Chain. The frog (f) is trying to capture the hopper (h). The only NE in the game contains non-uniformly random mixed policies.}
+        \includegraphics[width=0.931\textwidth]{images/biased_rps.png}
+        \caption{The initial state of Biased RPS in Food Chain. The frog (f) is trying to capture the hopper (h). The only NE in the game contains non-uniformly random mixed policies.}
         \label{fig:biased_rps_in_food_chain}
@@
         \includegraphics[width=0.9\textwidth]{images/tree_maze_chase.png}
-        \caption{Tree Maze Chase in Food Chain. The frog (f) is chasing the hooper (h) in the maze. There is at least one optimal pure policy for both players.}
+        \caption{The initial state of Tree Maze Chase in Food Chain. The frog (f) is chasing the hooper (h) in the maze. There is an optimal pure (deterministic) policy for both players.}
         \label{fig:tree_maze_chase}
@@
 \subsection{Biased Rock-Paper-Scissors in Food Chain}
-This scenario forces the agents to address the mixed strategy problem.
-We set up the environment as shown in Figure~\ref{fig:biased_rps_in_food_chain}.
-Since all agents have the same mobility, a single predator can only capture a prey by trapping them at a corner.
-All predators, unless working in a team, have to face this subcase at some point during a general Food Chain game.
-Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} in the appendix shows all possible next states.
+This case violates condition (\ref{eq:necessary_condition}) and forces the agents to address the mixed policy problem.
+The initial state of the scenario consists of a frog trying to hunt down a hopper as shown in Figure~\ref{fig:biased_rps_in_food_chain}.
 We can see that in some cases the hopper can escape.
 If they both pass, they will face the exact same situation again.
+Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} in the appendix shows all possible next states.
 For simplicity, we set the max timestep to 1, so the game ends after one move.
@@
 We also raise the feeding reward for both agents to 0.5 to highlight the asymmetry.
-Let hopper be the row player.
 The reward matrix is shown in Table~\ref{table:biased_rps_in_food_chain_reward_matrix} in the appendix.
-
 We can verify that there is a unique NE in the game.
@@
 
+In the Food Chain game, because all agents have the same mobility, a single predator can only capture a prey by trapping them in a corner.
+Thus, any predator, unless working in a team, has to face this subcase at some point during a general Food Chain game played on an open ground.
+This means that the scenario represents a subgame which will show up frequently in a full game.
+
 \subsection{Tree Maze Chase in Food Chain}
-This scenario essentially turns off the mixed strategy problem, so the condition (\ref{eq:necessary_condition}) is met. We set up the game environment as follows.
-
-This map setup resembles a tree structure (no cycles).
-It's easy to see that there is a pure NE in the game, because there is always an optimal action for both players.
-The agents don't need to adopt mixed policies.
+This scenario essentially turns off the mixed policy problem, so condition (\ref{eq:necessary_condition}) is met.
+We set up the initial state of the game as shown in Figure~\ref{fig:tree_maze_chase} where a frog is chasing a hopper in a maze.
+This map resembles a tree structure (no cycles) which ensures that there is a pure NE in the game, because there is always an optimal action for both players.
 The hooper should go for the tunnel with more food and the frog should simply chase the hopper.
-This means the necessary condition (\ref{eq:necessary_condition}) for IVBL to converge is met.
+This means the necessary condition (\ref{eq:necessary_condition}) is satisfied.
+To make sure that at most two food can be picked up by the hopper, we set the max timestep to 12.
+The optimal rewards for the hopper and the frog are -0.8 and 1, respectively.
 
@@
 In both scenarios, we test the IVBL algorithm IQL and the independent policy-based algorithm IPPO.
-The algorithm collects 640 timesteps and train the neural network in each iteration.
+All experiments were run for 800 iterations.
+For each iteration, the algorithm collects 640 timesteps as training samples and train the neural network.
 After each training iteration, 50 evaluation games are run during the evaluation phase.
@@
 In total, each data point on the graph is the average result of $20 \times 50 = 1000$ evaluation games.
+More details of the parameters for the experiments can be found in Table~\ref{table:experiments_parameters} in the appendix.
 
 Figure~\ref{fig:biased_rps_in_food_chain_rewards} demonstrates the performance of the algorithms in the biased RPS scenario where the necessary condition (\ref{eq:necessary_condition}) for IVBL to converge is not met.
-The results show that the IVBL algorithm IQL failed to converge to anything and are struggling to stabilize itself.
-The hopper's reward fluctuates between -1 and 0, and the frog reward oscillates in the range [0.1, 0.5].
-These values can be 100\% away from what it should get (-0.538 and 0.333).
+The results show that the IVBL algorithm IQL failed to converge to a defined solution range and were struggling to stabilize itself.
+The hopper's reward fluctuates between -1 and -0.2, where the optimal value is -0.538.
+These value can be as far as 85\% away from the desired amount.
+The frog reward oscillates in the range [0.2, 0.5], which is in a $\pm 70\%$  range from the actual NE (0.333).
 For comparison, the independent policy-based learning algorithm IPPO achieves a reward that stays more in the optimal range and is far more stable which indicates the agents are converging towards the NE.
@@
 Although still unstable, IQL can roughly reach the optimal reward at the end.
-The optimal reward for the hopper and the frog are -0.8 and 0.1, respectively.
+The optimal rewards for the hopper and the frog are -0.8 and 0.1, respectively.
 Even in this case, the policy-based method still appears to be more reliable than IVBL.
@@
 \subfloat[Biased RPS Hopper Rewards]{
-    \resizebox*{0.45\textwidth}{!}{
+    \resizebox*{0.46\textwidth}{!}{
         \includesvg{plots/food_chain_biased_rps_hopper_return.svg}
@@
 \subfloat[Biased RPS Frog Rewards]{
-    \resizebox*{0.45\textwidth}{!}{
+    \resizebox*{0.46\textwidth}{!}{
         \includesvg{plots/food_chain_biased_rps_frog_return.svg}
@@
 }
-\caption{Biased RPS Rewards}
+\caption{Biased RPS. IVBL (IQL) shows significantly more instability and does not converge to the NE.}
 \label{fig:biased_rps_in_food_chain_rewards}
@@
 \centering
-\subfloat[Biased RPS Hopper Rewards]{
-    \resizebox*{0.45\textwidth}{!}{
+\subfloat[Tree Maze Chase Hopper Rewards]{
+    \resizebox*{0.46\textwidth}{!}{
         \includesvg{plots/food_chain_tree_maze_chase_hopper_return.svg}
@@
 }
-\subfloat[Biased RPS Frog Rewards]{
-    \resizebox*{0.45\textwidth}{!}{
+\subfloat[Tree Maze Chase Frog Rewards]{
+    \resizebox*{0.46\textwidth}{!}{
         \includesvg{plots/food_chain_tree_maze_chase_frog_return.svg}
@@
 }
-\caption{Tree Maze Chase Rewards}
+\caption{Tree Maze Chase.
+Both algorithms quickly learned to play the game and tend to stay in the optimal reward range.
+Note that increasing frog reward reduces the hopper reward as the game is non-cooperative.}
 \label{fig:tree_maze_chase_rewards}
@@
 \section{Conclusion}
-We have identified a necessary condition (\ref{eq:necessary_condition}) for IVBL algorithms to converge to a NE in finite games.
+In this work, we have analysed the stability of IVBL algorithms in non-cooperative games and tested their performance in specific game scenarios.
+A necessary condition (\ref{eq:necessary_condition}) for IVBL algorithms to converge to a NE in finite games was identified.
+The condition requires the game to have a NE where every agent's policy evenly chooses their worthy actions (actions that are played with a positive probability).
+This explains why we may observe satisfying results when applying IVBL to games with pure NEs including cooperative games.
 We further proposed the game Food Chain which has certain traits that can verify our analysis.
-The results show that IVBL algorithms perform poorly in scenarios where the condition (\ref{eq:necessary_condition}) is violated.
+The test results show that IVBL algorithms perform poorly in scenarios where the condition (\ref{eq:necessary_condition}) is violated.
+Our work is limited to small two-player scenarios in the Food Chain game.
+There are still interesting characteristics of the game that may be explored in the future.
+Such as the extension in both the number of agents and the number of species in the food web.
 
 \section*{Acknowledgements}
-We thank the creators of the MARL library BenchMARL \cite{bettini2024benchmarl}, which was used to run the experiments.
+The authors sincerely gratitude the creators of the MARL library BenchMARL \cite{bettini2024benchmarl}, which was used to run the experiments.
 
@@
 }
+\tabnote{Hopper is the row player.}
 \label{table:biased_rps_in_food_chain_reward_matrix}
@@
 
+\section{Experiments Parameters}
+\begin{table}
+\tbl{Experiments Parameters}{
+    \begin{tabular}{lcc}
+        \toprule
+        \multicolumn{3}{l}{game parameters} \\
+        \midrule
+        & Biased RPS & Tree Maze Chase \\
+        \cmidrule{2-3}
+        size of the map & $2 \times 2$ & $5 \times 5$ \\
+        max timestep & 1 & 12 \\
+        feeding reward & 0.5 & 0.1 \\
+        death reward & \multicolumn{2}{c}{-1} \\
+        food spawn chance & \multicolumn{2}{c}{0} \\
+        number of agents & \multicolumn{2}{c}{2} \\
+        agent types & \multicolumn{2}{c}{2} \\
+        \midrule
+        \multicolumn{3}{l}{experiment parameters} \\
+        \midrule
+        repetitions of each experiment & \multicolumn{2}{c}{20} \\
+        number of iterations & \multicolumn{2}{c}{800} \\
+        evaluation games after each iteration & \multicolumn{2}{c}{50} \\
+        \midrule
+        \multicolumn{3}{l}{training hyper parameters} \\
+        \midrule
+        & IPPO & IQL \\
+        \cmidrule{2-3}
+        on-policy minibatch optimization iterations\textsuperscript{a} & 10 & \\
+        off-policy optimizer steps\textsuperscript{b} & & 100 \\
+        off-policy memory buffer size & & 20,000 \\
+        frames collected for each iteration & \multicolumn{2}{c}{640} \\
+        minibatch size & \multicolumn{2}{c}{32} \\
+        discount factor $\gamma$ & \multicolumn{2}{c}{0.99} \\
+        learning rate $\alpha$ & \multicolumn{2}{c}{$5 \times 10^{-5}$} \\
+        the $\varepsilon$ parameter of the adam optimizer & \multicolumn{2}{c}{$1 \times 10^{-6}$} \\
+        \bottomrule
+    \end{tabular}
+}
+\tabnote{\textsuperscript{a}The number of times collected frames will be split into minibatches and trained.
+\\ \textsuperscript{b}The number of times a minibatch will be sampled from the memory buffer and trained over.}
+\label{table:experiments_parameters}
+\end{table}
+
 \end{document}

### 2025-06-07 paper/paper.tex
paper v8

@@
 \begin{abstract}
-Multi-agent reinforcement learning has received a lot of updates in cooperative games recently.
+Multi-agent reinforcement learning has received increased attention in cooperative games recently.
 However, research in general non-cooperative games is currently lagging behind.
-Independent value-based algorithms such as independent Q-learning have been one of the most widely studied methods in the field due to their simplicity and scalability.
-Some research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios.
-In this paper, we discuss the feasibility of these algorithms in non-cooperative settings.
+Independent value-based algorithms, such as independent Q-learning, have demonstrated simple implementation and versatility in various contexts.
+Previous research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios.
+In this paper, we study the behaviour of these algorithms in non-cooperative settings.
 As a result, we explain the conditions that a game must satisfy for the algorithms to work.
 We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our analysis.
-Our results show that independent value-based learning algorithms are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
+We collect empirical data on the performance of selected algorithms from experiments based on the proposed game.
+Our results show that independent value-based learning algorithms
+can converge to Nash equilibrium in non-cooperative environments, only when the Nash equilibrium consists of uniformly random policies over the feasible actions.
+%are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
 \end{abstract}
@@
 The latter approach emphasis the action with higher empirical return.
-However, these approaches have a huge problem that guarantees failure of learning in a wide range of games.
+However, these approaches have a problem that guarantees failure of learning in a wide range of games.
 We will take a closer look in the analysis section.
@@
 \subfloat[A Food Web]{
-    \resizebox*{0.5\textwidth}{!}{
+    \resizebox*{0.48\textwidth}{!}{
         \includegraphics{images/food_web.png}
@@
 \subfloat[The Food Chain Game]{
-    \resizebox*{0.5\textwidth}{!}{
+    \resizebox*{0.48\textwidth}{!}{
         \includegraphics{images/food_chain_game.png}
@@
 The arrows point in the direction of the energy flow.
-(b), In the food chain game, an agent represents an individual animal of some type.
+(b), In the Food Chain game, an agent represents an individual animal of some type.
 They navigate on a playground trying to feed themselves while avoiding enemies.}
@@
 Hoppers (represented by h) can consume vegetation (drawn as *) and will be eaten by frogs (f).
-Both scenarios will turn off random food generation.
+Both scenarios turns off vegetation generation.
 
@@
         \includegraphics[width=0.9\textwidth]{images/tree_maze_chase.png}
-        \caption{The initial state of Tree Maze Chase in Food Chain. The frog (f) is chasing the hooper (h) in the maze. There is an optimal pure (deterministic) policy for both players.}
+        \caption{The initial state of Tree Maze Chase in Food Chain. The frog (f) is chasing the hopper (h) in the maze. There is an optimal pure (deterministic) policy for both players.}
         \label{fig:tree_maze_chase}
@@
 \subsection{Biased Rock-Paper-Scissors in Food Chain}
-This case violates condition (\ref{eq:necessary_condition}) and forces the agents to address the mixed policy problem.
+This case violates condition (\ref{eq:necessary_condition}).
 The initial state of the scenario consists of a frog trying to hunt down a hopper as shown in Figure~\ref{fig:biased_rps_in_food_chain}.
@@
 Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} in the appendix shows all possible next states.
-For simplicity, we set the max timestep to 1, so the game ends after one move.
-The scenario becomes a normal form mini game.
-To break the symmetry so that condition (\ref{eq:necessary_condition}) is violated, we add food in one cell, making the scenario essentially biased RPS.
+We set the max timestep to 1, so the scenario becomes a normal form game.
+To break the symmetry so that condition (\ref{eq:necessary_condition}) is violated, we add vegetation in one cell, making the scenario essentially biased RPS.
 We also raise the feeding reward for both agents to 0.5 to highlight the asymmetry.
@@
 \end{equation}
-The expected reward for the hooper and the frog should be $-7/13 \approx -0.538$ and $1/3 \approx 0.333$, respectively.
+The expected reward for the hopper and the frog should be $-7/13 \approx -0.538$ and $1/3 \approx 0.333$, respectively.
 
-In the Food Chain game, because all agents have the same mobility, a single predator can only capture a prey by trapping them in a corner.
+In the Food Chain game, because all agents have the same mobility, a predator can only capture a prey by trapping them in a corner.
 Thus, any predator, unless working in a team, has to face this subcase at some point during a general Food Chain game played on an open ground.
-This means that the scenario represents a subgame which will show up frequently in a full game.
+The scenario represents a subgame which shows up frequently in a full game.
 
 \subsection{Tree Maze Chase in Food Chain}
-This scenario essentially turns off the mixed policy problem, so condition (\ref{eq:necessary_condition}) is met.
-We set up the initial state of the game as shown in Figure~\ref{fig:tree_maze_chase} where a frog is chasing a hopper in a maze.
-This map resembles a tree structure (no cycles) which ensures that there is a pure NE in the game, because there is always an optimal action for both players.
-The hooper should go for the tunnel with more food and the frog should simply chase the hopper.
-This means the necessary condition (\ref{eq:necessary_condition}) is satisfied.
-To make sure that at most two food can be picked up by the hopper, we set the max timestep to 12.
+This scenario essentially turns off the mixed policy problem.
+We set up the initial state as shown in Figure~\ref{fig:tree_maze_chase} where a frog is chasing a hopper in a maze.
+This map resembles a tree structure (no cycles) which creates a pure NE, because there is always an optimal action.
+The hopper should go for the tunnel with more vegetation and the frog should simply chase the hopper.
+The necessary condition (\ref{eq:necessary_condition}) is satisfied.
+To make sure that at most two vegetations can be picked up by the hopper, we set the max timestep to 12.
 The optimal rewards for the hopper and the frog are -0.8 and 1, respectively.
@@
 \subsection{Results and Discussion}
-In both scenarios, we test the IVBL algorithm IQL and the independent policy-based algorithm IPPO.
-All experiments were run for 800 iterations.
-For each iteration, the algorithm collects 640 timesteps as training samples and train the neural network.
-After each training iteration, 50 evaluation games are run during the evaluation phase.
+We tested the IVBL algorithm IQL and the independent policy-based algorithm IPPO in both scenarios.
+All experiments ran for 800 iterations.
+For each iteration, the algorithm collects 640 timesteps as training samples and train the neural network before 50 evaluation games were run.
 The plots show the average reward for each agent during the evaluation phase.
-Each experiment is run 20 times to mitigate the effect of randomness and the graphs show the boots trapped 95\% confidence interval.
-In total, each data point on the graph is the average result of $20 \times 50 = 1000$ evaluation games.
+Each experiment was run 20 times to mitigate the effect of randomness and the graphs show the boots trapped 95\% confidence interval.
 More details of the parameters for the experiments can be found in Table~\ref{table:experiments_parameters} in the appendix.
+
+\begin{figure}
+\centering
+\subfloat[Biased RPS Hopper Rewards]{
+    \resizebox*{0.48\textwidth}{!}{
+        \includesvg{plots/food_chain_biased_rps_hopper_return.svg}
+    }
+}
+\subfloat[Biased RPS Frog Rewards]{
+    \resizebox*{0.48\textwidth}{!}{
+        \includesvg{plots/food_chain_biased_rps_frog_return.svg}
+    }
+}
+\caption{Biased RPS. IVBL (IQL) shows significantly more instability and does not converge to the NE.}
+\label{fig:biased_rps_in_food_chain_rewards}
+\end{figure}
+
+\begin{figure}
+\centering
+\subfloat[Tree Maze Chase Hopper Rewards]{
+    \resizebox*{0.48\textwidth}{!}{
+        \includesvg{plots/food_chain_tree_maze_chase_hopper_return.svg}
+    }
+}
+\subfloat[Tree Maze Chase Frog Rewards]{
+    \resizebox*{0.48\textwidth}{!}{
+        \includesvg{plots/food_chain_tree_maze_chase_frog_return.svg}
+    }
+}
+\caption{Tree Maze Chase.
+Both algorithms quickly learned to play the game and tend to stay in the optimal reward range.
+Note that increasing frog reward reduces the hopper reward as the game is non-cooperative.}
+\label{fig:tree_maze_chase_rewards}
+\end{figure}
 
@@
 
-\begin{figure}
-\centering
-\subfloat[Biased RPS Hopper Rewards]{
-    \resizebox*{0.46\textwidth}{!}{
-        \includesvg{plots/food_chain_biased_rps_hopper_return.svg}
-    }
-}
-\subfloat[Biased RPS Frog Rewards]{
-    \resizebox*{0.46\textwidth}{!}{
-        \includesvg{plots/food_chain_biased_rps_frog_return.svg}
-    }
-}
-\caption{Biased RPS. IVBL (IQL) shows significantly more instability and does not converge to the NE.}
-\label{fig:biased_rps_in_food_chain_rewards}
-\end{figure}
+\section{Conclusion}
+In this work, we have analysed the stability of independent value-based learning (IBVL) algorithms in non-cooperative games and tested their performance in specific game scenarios.
+By means of proof by contradiction and the design of experiments based on the proposed Food Chain game, we found that the studied algorithms only converge to Nash equilibrium (NE) in non-cooperative settings when every agent's policy evenly chooses the next action among their worthy actions (actions that are played with a positive probability).
+This explains why promising results are achieved when applying IVBL to games with pure NE, including cooperative games.
 
-\begin{figure}
-\centering
-\subfloat[Tree Maze Chase Hopper Rewards]{
-    \resizebox*{0.46\textwidth}{!}{
-        \includesvg{plots/food_chain_tree_maze_chase_hopper_return.svg}
-    }
-}
-\subfloat[Tree Maze Chase Frog Rewards]{
-    \resizebox*{0.46\textwidth}{!}{
-        \includesvg{plots/food_chain_tree_maze_chase_frog_return.svg}
-    }
-}
-\caption{Tree Maze Chase.
-Both algorithms quickly learned to play the game and tend to stay in the optimal reward range.
-Note that increasing frog reward reduces the hopper reward as the game is non-cooperative.}
-\label{fig:tree_maze_chase_rewards}
-\end{figure}
-
-\section{Conclusion}
-In this work, we have analysed the stability of IVBL algorithms in non-cooperative games and tested their performance in specific game scenarios.
-A necessary condition (\ref{eq:necessary_condition}) for IVBL algorithms to converge to a NE in finite games was identified.
-The condition requires the game to have a NE where every agent's policy evenly chooses their worthy actions (actions that are played with a positive probability).
-This explains why we may observe satisfying results when applying IVBL to games with pure NEs including cooperative games.
-We further proposed the game Food Chain which has certain traits that can verify our analysis.
-The test results show that IVBL algorithms perform poorly in scenarios where the condition (\ref{eq:necessary_condition}) is violated.
+This work also proposes the Food Chain game which has certain traits ---such as ease of navigation, focus on interactions, and extendability, among others--- that are instrumental for the analysis of the stability in non-cooperative settings.
+The test results show that IVBL algorithms perform poorly in scenarios where the identified condition is violated.
 Our work is limited to small two-player scenarios in the Food Chain game.
-There are still interesting characteristics of the game that may be explored in the future.
-Such as the extension in both the number of agents and the number of species in the food web.
+There are still interesting characteristics of the game that will be explored in the future.
+Such as the extension in both the number of agents and the number of species on the food web.
 
@@
 Figure~\ref{fig:biased_rps_in_food_chain_possible_next_states} shows all the possible next state of the Biased RPS scenario in Food Chain.
-Table~\ref{table:biased_rps_in_food_chain_reward_matrix} describes the reward matrix of the scenario (note that if both agents try to move into the food tile, the hopper only has a 50\% chance to get food).
+Table~\ref{table:biased_rps_in_food_chain_reward_matrix} describes the reward matrix of the scenario (note that if both agents try to move into the vegetation tile, the hopper only has a 50\% chance to get the vegetation).
 
@@
         death reward & \multicolumn{2}{c}{-1} \\
-        food spawn chance & \multicolumn{2}{c}{0} \\
+        vegetation spawn chance & \multicolumn{2}{c}{0} \\
         number of agents & \multicolumn{2}{c}{2} \\

### 2025-06-07 research/paper/paper.tex
added data folder

--- 2025-06-07 paper/paper.tex
@@
 \begin{abstract}
-Multi-agent reinforcement learning has received increased attention in cooperative games recently.
-However, research in general non-cooperative games is currently lagging behind.
-Independent value-based algorithms, such as independent Q-learning, have demonstrated simple implementation and versatility in various contexts.
-Previous research has shown that independent algorithms can heavily outperform other approaches in some cooperative scenarios.
+Multi-agent reinforcement learning has received increased attention in cooperative games.
+However, research in non-cooperative games is lagging behind.
+Independent value-based learning algorithms have demonstrated simplicity and versatility in various contexts.
 In this paper, we study the behaviour of these algorithms in non-cooperative settings.
-As a result, we explain the conditions that a game must satisfy for the algorithms to work.
-We further propose the game Food Chain that simulates an ecosystem and is designed to challenge the algorithms in a way aligned with our analysis.
-We collect empirical data on the performance of selected algorithms from experiments based on the proposed game.
-Our results show that independent value-based learning algorithms
-can converge to Nash equilibrium in non-cooperative environments, only when the Nash equilibrium consists of uniformly random policies over the feasible actions.
-%are guaranteed to fail in some non-cooperative games but are more likely to work in cooperative games.
+We explain the conditions that a game must satisfy for the algorithms to work.
+We further test the algorithms in our proposed game Food Chain that simulates an ecosystem.
+Our results show that independent value-based learning algorithms can converge to Nash equilibrium, only when the Nash equilibrium consists of uniformly random policies over the feasible actions.
 \end{abstract}
@@
 \begin{keywords}
-Multi-agent; reinforcement learning; independent value-based learning; game theory; non-cooperative games; competitive games; Nash equilibrium
+Multi-agent; reinforcement learning; independent value-based learning; game theory; non-cooperative games; competitive games; Nash equilibrium; food chain; food web; ecosystem
 \end{keywords}
@@
 They can move around the map and feed themselves while trying to avoid enemies.
-Example game scenarios are shown in Figure~\ref{fig:biased_rps_in_food_chain} and Figure~\ref{fig:tree_maze_chase}.
 The game simulates an ecosystem to a certain extend, by defining a food web structure.
@@
 \centering
-\subfloat[A Food Web]{
+\subfloat[A Food Web\textsuperscript{1}]{
     \resizebox*{0.48\textwidth}{!}{
@@
 }
-\subfloat[The Food Chain Game]{
+\subfloat[The Food Chain Game\textsuperscript{1}]{
     \resizebox*{0.48\textwidth}{!}{
@@
 }
-\caption{(a), A food web represents a simple ecosystem.
+\caption[Caption for LOF]{(a), A food web represents a simple ecosystem.
 The food web is a cluster of multiple food chains showing what-eats-what.
@@
 \end{figure}
+\footnotetext[1]{The individual animal avatars were generated by ChatGPT GPT-4-turbo model which includes image generation via DALL·E 3, provided by OpenAI.}
 

### 2025-06-07 research/paper/paper.tex
data
less key words

@@
 \begin{keywords}
-Multi-agent; reinforcement learning; independent value-based learning; game theory; non-cooperative games; competitive games; Nash equilibrium; food chain; food web; ecosystem
+Multi-agent reinforcement learning; independent value-based learning; game theory; non-cooperative games; Nash equilibrium
 \end{keywords}
@@
 We tested the IVBL algorithm IQL and the independent policy-based algorithm IPPO in both scenarios.
+The experiments were run using the open-source MARL library BenchMARL \cite{bettini2024benchmarl}.
 All experiments ran for 800 iterations.
@@
 Such as the extension in both the number of agents and the number of species on the food web.
-
-\section*{Acknowledgements}
-The authors sincerely gratitude the creators of the MARL library BenchMARL \cite{bettini2024benchmarl}, which was used to run the experiments.
 
